### 2015-10-10 doc/template.tex
Primeira versão do relatório

\documentclass[a4paper]{article}

%use the english line for english reports
%usepackage[english]{babel}
\usepackage[portuguese]{babel}
\usepackage[utf8]{inputenc}
\usepackage{indentfirst}
\usepackage{graphicx}
\usepackage{verbatim}

\begin{document}

\setlength{\textwidth}{16cm}
\setlength{\textheight}{22cm}

\title{\Huge\textbf{Breakthru}\linebreak\linebreak\linebreak
\Large\textbf{Relatório Intercalar}\linebreak\linebreak
\linebreak\linebreak
\includegraphics[scale=0.1]{feup-logo.png}\linebreak\linebreak
\linebreak\linebreak
\Large{Mestrado Integrado em Engenharia Informática e Computação} \linebreak\linebreak
\Large{Programação em Lógica}\linebreak
}

\author{\textbf{Grupo xx:}\\
João Rafael de Figueiredo Cabral - 201304395 \\
João Bernardo Martins de Sousa e Silva Mota - Número2 \\
\linebreak\linebreak \\
 \\ Faculdade de Engenharia da Universidade do Porto \\ Rua Roberto Frias, s\/n, 4200-465 Porto, Portugal \linebreak\linebreak\linebreak
\linebreak\linebreak\vspace{1cm}}

\maketitle
\thispagestyle{empty}

%************************************************************************************************
%************************************************************************************************

\newpage

%Todas as figuras devem ser referidas no texto. %\ref{fig:codigoFigura}
%
%%Exemplo de código para inserção de figuras
%%\begin{figure}[h!]
%%\begin{center}
%%escolher entre uma das seguintes três linhas:
%%\includegraphics[height=20cm,width=15cm]{path relativo da imagem}
%%\includegraphics[scale=0.5]{path relativo da imagem}
%%\includegraphics{path relativo da imagem}
%%\caption{legenda da figura}
%%\label{fig:codigoFigura}
%%\end{center}
%%\end{figure}
%
%
%\textit{Para escrever em itálico}
%\textbf{Para escrever em negrito}
%Para escrever em letra normal
%``Para escrever texto entre aspas''
%
%Para fazer parágrafo, deixar uma linha em branco.
%
%Como fazer bullet points:
%\begin{itemize}
	%\item Item1
	%\item Item2
%\end{itemize}
%
%Como enumerar itens:
%\begin{enumerate}
	%\item Item 1
	%\item Item 2
%\end{enumerate}
%
%\begin{quote}``Isto é uma citação''\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{O Jogo Breakthru}
O \emph{Breakthru} é um jogo de estratégia abstrata para dois jogadores em que cada lado começa com um número diferente de peças, colocadas em posições assimétricas e com objetivos diferentes para assegurar a vitória.

\subsection{História}
O jogo foi criado em 1965 por Alex Randolph, com um conjunto de regras baseados nas do \emph{Hnefatafl}\footnote{https://boardgamegeek.com/boardgame/335/breakthru}, um jogo de estratégia abstrata cuja popularidade remonta à Escandinávia medieval, existindo vestigios da sua presença entre várias civilizações antigas\footnote{https://boardgamegeek.com/boardgame/2932/hnefatafl}.
\

\subsection{Regras}
\subsubsection{Peças}
A cor de cada jogador é sorteada através do lançamento de uma moeda, sendo que o jogador vencedor desse sorteio é o jogador dourado e o perdedor é o jogador prateado. O tabuleiro conta com 13 peças douradas, das quais 12 dispostas em quadrado à volta da 13ª peça, no centro do tabuleiro, e de 20 peças prateadas dispostas em volta das peças douradas.

O jogador dourado dispõe as suas 12 peças normais como desejar na zona central do tabuleiro, nele delimitada. O jogador prateado procede de forma semelhante, dispondo contudo as suas 20 peças fora da zona central, reservada ao jogador dourado.

Uma das peças douradas é diferente das restantes, sendo designada por navio almirante. A sua posição inicial é fixa, sendo obrigatório que seja colocada no centro do tabuleiro.

\subsubsection{Movimentos e Capturas}
Cada jogador deve, na sua vez, efetuar dois movimentos ou uma captura.

Todas as peças podem ser movimentadas um número arbitrário de espaços livres na horizontal ou vertical, a semelhança do movimento de uma torre no xadrez. Durante o movimento nenhuma peça pode saltar por cima de outra ou ocupar uma casa já ocupada, como ilustrado na figura \ref{fig:moves}.

As capturas são feitas movimentando uma peça para uma casas que lhe seja diagonalmente adjacente e que esteja ocupada por uma peça do adversário, como ilustrado na figura \ref{fig:captures}, sendo esta removida do jogo.

Quando o navio almirante dourado efetua um movimento o jogador dourado não pode efetuar outro movimento nesse turno.

\subsubsection{Objetivo do Jogo}
O jogador prateado ganha o jogo ao capturar o navio almirante dourado. O jogador dourado ganha o jogo se o navio almirante chegar a uma das casas exteriores do tabuleiro.

\begin{figure}
\centering
\includegraphics[scale=1]{Breakthru_board.png} 
\centering 
\caption{Uma posição inicial possível das peças num tabuleiro de Breakthru}
\label{fig:init}
\end{figure}
\begin{figure}
\centering
\includegraphics[scale=1]{Breakthru_moves.png}
\caption{A vermelho: movimentos possíveis da peça dourada}
\label{fig:captures}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=1]{Breakthru_capture.png}
\caption{A vermelho: capturas possíveis da peça dourada}
\label{fig:moves}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Representação do Estado do Jogo}

Descrever a forma de representação do estado do tabuleiro (tipicamente uma lista de listas), com exemplificação em Prolog de posições iniciais do jogo, posições intermédias e finais, acompanhadas de imagens ilustrativas.

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Visualização do Tabuleiro}

Descrever a forma de visualização do tabuleiro em modo de texto e o(s) predicado(s) Prolog construídos para o efeito.
Deve ser incluída pelo menos uma imagem correspondente ao output produzido pelo predicado de visualização.

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Movimentos}

Elencar os movimentos (tipos de jogadas) possíveis e definir os cabeçalhos dos predicados que serão utilizados (ainda não precisam de estar implementados).

\end{document}

### 2015-10-11 board.pl
Fiz alguns bonitos na impressão do tabuleiro
E criei tabuleiros de exemplo

@@ -1,29 +1,65 @@
 
 %create predicates based on ferrolhos implementation
+exampleInitial([ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+                 [' ',' ',' ','p','p','p','p','p',' ',' ',' '],
+                 [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+                 [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
+                 [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
+                 [' ','p',' ','d',' ','F',' ','d',' ','p',' '],
+                 [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
+                 [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
+                 [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+                 [' ',' ','p','p','p','p','p','p',' ',' ',' '],
+                 [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '] ]).
 
-createBoard(_, 0, L).
+exampleMiddle([ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+                [' ',' ',' ','p','p',' ','d','p',' ',' ',' '],
+                [' ',' ',' ','d',' ',' ',' ',' ',' ',' ',' '],
+                [' ',' ',' ','p','d',' ','d','p',' ',' ',' '],
+                [' ','p',' ',' ',' ','p',' ','d',' ','p',' '],
+                [' ','p',' ','d',' ','F',' ','d',' ','p',' '],
+                [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
+                [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
+                [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+                [' ',' ','p','p','p','p','p','p',' ',' ',' '],
+                [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '] ]).
+
+exampleEnd([ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+                [' ',' ',' ','p','p',' ','d','p',' ',' ',' '],
+                [' ',' ',' ','d',' ',' ',' ',' ',' ',' ',' '],
+                [' ',' ',' ','p','d',' ','d','p',' ',' ',' '],
[…]

### 2015-10-11 doc/template.tex
corrigidos numeros no relatorio
relatorio atualizado

@@
 
-\author{\textbf{Grupo xx:}\\
+\author{\textbf{Grupo Breakthru 2:}\\
 João Rafael de Figueiredo Cabral - 201304395 \\
-João Bernardo Martins de Sousa e Silva Mota - Número2 \\
+João Bernardo Martins de Sousa e Silva Mota - 201303462 \\
 \linebreak\linebreak \\
@@
 \section{Representação do Estado do Jogo}
+O tabuleiro é representado como uma lista de listas. De seguida apresentam-se  exemplos de tabuleiros iniciais, intermédios e finais criados em Prolog.
 
-Descrever a forma de representação do estado do tabuleiro (tipicamente uma lista de listas), com exemplificação em Prolog de posições iniciais do jogo, posições intermédias e finais, acompanhadas de imagens ilustrativas.
+Possível estado inicial:
+
+\begin{verbatim} 
+[[' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+ [' ',' ',' ','p','p','p','p','p',' ',' ',' '],
+ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+ [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
+ [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
+ [' ','p',' ','d',' ','F',' ','d',' ','p',' '],
+ [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
+ [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
+ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+ [' ',' ','p','p','p','p','p','p',' ',' ',' '],
+ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '] ] 
+\end{verbatim}
+
+Possível estado intermédio:
+\begin{verbatim}
+[[' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+ [' ',' ',' ','p','p',' ','d','p',' ',' ',' '],
+ [' ',' ',' ','d',' ',' ',' ',' ',' ',' ',' '],
+ [' ',' ',' ','p','d',' ','d','p',' ',' ',' '],
+ [' ','p',' ',' ',' ','p',' ','d',' ','p',' '],
+ [' ','p',' ','d',' ','F',' ','d',' ','p',' '],
+ [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
+ [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
+ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+ [' ',' ','p','p','p','p','p','p',' ',' ',' '],
+ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '] ]
+\end{verbatim}
+
+Possível estado final, onde as peças douradas venceram porque o navio almirante atingiu uma das casas exteriores:
+\begin{verbatim}
+[[' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+ [' ',' ',' ','p','p',' ','d','p',' ',' ',' '],
+ [' ',' ',' ','d',' ',' ',' ',' ',' ',' ',' '],
+ [' ',' ',' ','p','d',' ','d','p',' ',' ',' '],
+ [' ','p',' ',' ',' ','p',' ','d',' ','p',' '],
+ [' ','p',' ','d',' ',' ',' ',' ','d','p',' '],
+ [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
+ [' ','p',' ','p','d',' ','d',' ',' ','p',' '],
+ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
+ [' ',' ','p','','d',' ','p','p',' ',' ',' '],
+ [' ',' ',' ',' ',' ','F',' ',' ',' ',' ',' '] ] 
+\end{verbatim}
 
@@
 
-Descrever a forma de visualização do tabuleiro em modo de texto e o(s) predicado(s) Prolog construídos para o efeito.
-Deve ser incluída pelo menos uma imagem correspondente ao output produzido pelo predicado de visualização.
+Foram desenvolvidos alguns predicados em Prolog para permitir a visualização do tabuleiro usando caracteres \textit{ASCII}, como exemplificado pela figura \ref{fig:example}.
+\begin{figure}
+\centering
+\includegraphics[scale=1]{Breakthru_initial_sicstus.png}
+\caption{Tabuleiro inicial de um jogo de Breakthru}
+\label{fig:example}
+\end{figure}
+
+\subsection{Predicados de Visualização}
+Foram implementados os predicados de visualização que se seguem:
+
+\begin{verbatim}
+printBoard([]):-write('----------------------------------------------').
+printBoard([H|T]):-
+  write('---------------------------------------------'),nl,
+  printLine(H, '|'), nl,
+  printBoard(T).
+
+printLine([], Sep):-
+  write(Sep).
+printLine([H|T], Sep):-
+  write(Sep),write(' '), write(H), write(' '),
+  printLine(T, Sep).
+\end{verbatim}
 
@@
 \section{Movimentos}
+São possíveis movimentos simples ou capturas, para os quais se desenvolverão os predicados cujos cabeçalhos se elencam na secção \ref{sec:predicates}, que devem suceder somente se o movimento for legal ou for possível efetuar a captura, respetivamente.
 
-Elencar os movimentos (tipos de jogadas) possíveis e definir os cabeçalhos dos predicados que serão utilizados (ainda não precisam de estar implementados).
+\subsection{Cabeçalhos dos Predicados} \label{sec:predicates}
+\begin{verbatim}
+movePiece(RowOrigin, ColumnOrigin, RowDestiny, ColumnDestiny,
+ Player, Board, NewBoard):-
+
+capturePiece(RowOrigin, ColumnOrigin, RowDestiny, ColumnDestiny,
+ Player, Board, NewBoard):-
+\end{verbatim}
 

### 2015-10-11 doc/template.tex
corrigidas gralhas

@@
 \subsection{História}
-O jogo foi criado em 1965 por Alex Randolph, com um conjunto de regras baseados nas do \emph{Hnefatafl}\footnote{https://boardgamegeek.com/boardgame/335/breakthru}, um jogo de estratégia abstrata cuja popularidade remonta à Escandinávia medieval, existindo vestigios da sua presença entre várias civilizações antigas\footnote{https://boardgamegeek.com/boardgame/2932/hnefatafl}.
+O jogo foi criado em 1965 por Alex Randolph, com um conjunto de regras baseadas nas do \emph{Hnefatafl}\footnote{https://boardgamegeek.com/boardgame/335/breakthru}, um jogo de estratégia abstrata cuja popularidade remonta à Escandinávia medieval, existindo vestígios da sua presença entre várias civilizações antigas\footnote{https://boardgamegeek.com/boardgame/2932/hnefatafl}.
 \
@@
 
-Todas as peças podem ser movimentadas um número arbitrário de espaços livres na horizontal ou vertical, a semelhança do movimento de uma torre no xadrez.
+Todas as peças podem ser movimentadas um número arbitrário de espaços livres na horizontal ou vertical, à semelhança do movimento de uma torre no xadrez.
 Durante o movimento nenhuma peça pode saltar por cima de outra ou ocupar uma casa já ocupada, como ilustrado na figura \ref{fig:moves}.
 
-As capturas são feitas movimentando uma peça para uma casas que lhe seja diagonalmente adjacente e que esteja ocupada por uma peça do adversário, como ilustrado na figura \ref{fig:captures}, sendo esta removida do jogo.
+As capturas são feitas movimentando uma peça para uma casa que lhe seja diagonalmente adjacente e que esteja ocupada por uma peça do adversário, como ilustrado na figura \ref{fig:captures}, sendo esta removida do jogo.
 

### 2015-10-13 board.pl
fixed replaceinBoard. added some files.

@@ -1,5 +1,4 @@
 
-%create predicates based on ferrolhos implementation
 exampleInitial([ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
                  [' ',' ',' ','p','p','p','p','p',' ',' ',' '],
                  [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
@@ -45,7 +44,7 @@ createBoard(N, M, [H|T]):-
   createBoard(N, M1, T).
 
 createLine(0, []).
-createLine(N, [H|T]):-
+createLine(N, [emptyCell|T]):-
   N > 0,
   N1 is N-1,
   createLine(N1, T).
@@ -63,16 +62,23 @@ printLine([H|T], Sep):-
   printLine(T, Sep).
 
 
-replaceInLine(0, Char, [Char|T]).
-replaceInLine(Y, Char, [H|T]):-
+replaceInLine(0, Char, [H|T], [Char|T]).
+replaceInLine(Y, Char, [H|T], [H|R]):-
   Y > 0,
   Y1 is Y-1,
-  replaceInLine(Y1, Char, T).
+  replaceInLine(Y1, Char, T, R).
 
-replaceInBoard(Y, 0, Char, [H|T]):-
-  replaceInLine(Y, Char, H).
+replaceInBoard(Y, 0, Char, [H|T], [L|T]):-
+  replaceInLine(Y, Char, H, L).
 
-replaceInBoard(Y, X, Char, [H|T]):-
+replaceInBoard(Y, X, Char, [H|T], [H|R]):-
   X > 0,
   X1 is X-1,
-  replaceInBoard(Y, X1, Char, T).
+  replaceInBoard(Y, X1, Char, T, R).
+
+  
+test:-
+	createBoard(8, 8, L),
+	printBoard(L), nl, nl,
+	replaceInBoard(4, 4, 'x', L, Nl),
+	printBoard(Nl).

### 2015-10-13 game.pl

@@ -1,0 +1,11 @@
+:- use_module(library(random)).
+:- include('board.pl').
+
+piece(goldenPiece).
+piece(silverPiece).
+piece(flagship).
+
+owner(goldenPlayer, goldenPiece).
+owner(silverPlayer, silverPiece).
+owner(goldenPlayer, flagship).
+

### 2015-10-13 util.pl

@@ -1,0 +1,12 @@
+use_module(library(random)).
+
+setSeed:-
+	now(Now),
+	Seed is Now mod 30269,
+	getrand(random(X, Y, Z, _)),
+	setrand(random(Seed, X, Y, Z)), 
+	!.
+
+coinToss(X):-
+	setSeed,
+	random(0, 1, X).

### 2015-10-13 board.pl
board imprime mai bonito

@@ -36,6 +36,11 @@ exampleEnd([ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
                 [' ',' ',' ',' ',' ','F',' ',' ',' ',' ',' '] ]).
 
 
+symbol(emptyCell, ' ').
+symbol(flagship, 'F').
+symbol(silverPiece, 'p').
+symbol(goldenPiece, 'd').
+				
 createBoard(_, 0, []).
 createBoard(N, M, [H|T]):-
   M > 0,
@@ -49,16 +54,21 @@ createLine(N, [emptyCell|T]):-
   N1 is N-1,
   createLine(N1, T).
 
-printBoard([]):-write('----------------------------------------------').
-printBoard([H|T]):-
-  write('---------------------------------------------'),nl,
-  printLine(H, '|'), nl,
-  printBoard(T).
+printBoard([], 0):-write(' ----------------------------------------------'),nl,
+	write('   0   1   2   3   4   5   6   7   8   9  10').
+	
+printBoard([H|T], N):-
+  N1 is N-1,
+  Print is 10-N1,
+  write(' ---------------------------------------------'),nl,
+  write(Print), printLine(H, '|'), nl,
+  printBoard(T, N1).
 
 printLine([], Sep):-
   write(Sep).
 printLine([H|T], Sep):-
-  write(Sep),write(' '), write(H), write(' '),
+  symbol(H, Symb),
+  write(Sep),write(' '), write(Symb), write(' '),
   printLine(T, Sep).
 
 

### 2015-10-13 game.pl

@@ -1,5 +1,6 @@
 :- use_module(library(random)).
 :- include('board.pl').
+:- include('util.pl').
 
 piece(goldenPiece).
 piece(silverPiece).
@@ -9,3 +10,9 @@ owner(goldenPlayer, goldenPiece).
 owner(silverPlayer, silverPiece).
 owner(goldenPlayer, flagship).
 
+startGame:- createBoard(11,11,BoardEmpty),
+			replaceInBoard(5,5,flagship, BoardEmpty, BoardFlag),
+			printBoard(BoardFlag,11).
+	
+	
+	

### 2015-10-13 util.pl

@@ -1,12 +1,2 @@
-use_module(library(random)).
-
-setSeed:-
-	now(Now),
-	Seed is Now mod 30269,
-	getrand(random(X, Y, Z, _)),
-	setrand(random(Seed, X, Y, Z)), 
-	!.
-
 coinToss(X):-
-	setSeed,
-	random(0, 1, X).
+	random(0, 2, X).

### 2015-11-05 board.pl
Move validation

@@ -40,7 +40,17 @@ symbol(emptyCell, ' ').
 symbol(flagship, 'F').
 symbol(silverPiece, 'p').
 symbol(goldenPiece, 'd').
-				
+
+getBoardElement([H|_], Piece, X, Y):- getElement(H, Piece).
+getBoardElement([_|T], Piece):- getBoardElement(T, Piece).
+getBoardElement([], _):- fail.
+
+getElement([H|_], H).
+getElement([_|T], Piece):- getElement(T, Piece).
+getElement([], _):- fail.
+
+
+
 createBoard(_, 0, []).
 createBoard(N, M, [H|T]):-
   M > 0,
@@ -56,7 +66,7 @@ createLine(N, [emptyCell|T]):-
 
 printBoard([], 0):-write(' ----------------------------------------------'),nl,
 	write('   0   1   2   3   4   5   6   7   8   9  10'), nl.
-	
+
 printBoard([H|T], N):-
   N1 is N-1,
   Print is 10-N1,
@@ -72,7 +82,7 @@ printLine([H|T], Sep):-
   printLine(T, Sep).
 
 
-replaceInLine(0, Char, [H|T], [Char|T]).
+replaceInLine(0, Char, [_|T], [Char|T]).
 replaceInLine(Y, Char, [H|T], [H|R]):-
   Y > 0,
   Y1 is Y-1,
@@ -86,15 +96,20 @@ replaceInBoard(Y, X, Char, [H|T], [H|R]):-
   X1 is X-1,
   replaceInBoard(Y, X1, Char, T, R).
 
-  
-getCell([H|T], X, 0, E):-
+
+getCell([H|_], X, 0, E):-
 	getCellL(H, X, E).
-getCell([H|T], X, Y, E):-
+getCell([_|T], X, Y, E):-
 	Y1 is Y-1,
 	getCell(T, X, Y1, E).
-	
+
 getCellL([H|_], 0, H).
-getCellL([H|T], X, E):-
+getCellL([_|T], X, E):-
 	X1 is X-1,
 	getCellL(T, X1, E).
-	
+
+fillBoardLine(11,Y).
+fillBoardLine(X,Y):-assertz(position(X,Y, emptyCell)), X1 is X+1, fillBoardLine(X1,Y).
+
+fillBoard(X,11).
+fillBoard(X,Y):-fillBoardLine(X,Y), Y1 is Y+1, fillBoard(X,Y1).

### 2015-11-05 game.pl

@@ -7,6 +7,7 @@
 :- dynamic(silverPieces/1).
 :- dynamic(goldenPieces/1).
 :- dynamic(gameState/2).
+:- dynamic(position/3).
 
 piece(goldenPiece).
 piece(silverPiece).
@@ -28,7 +29,7 @@ state(begin).
 state(game).
 state(over).
 
-teste(X):- createBoard(3,3,Board),
+teste(_):- createBoard(3,3,Board),
 		replaceInBoard(2,2, boda, Board, NewBoard),
 		getCell(NewBoard, 2, 2, boda).
 
@@ -51,6 +52,7 @@ playGame(Board):-
 
 
 setupBoard(Board, NewBoard):- assert(gameState(Board, goldenPlayer)),
+					fillBoard(11,11),
 					repeat,
 					retract(gameState(CurrentBoard, CurrentPlayer)),
 					write(CurrentPlayer),
@@ -73,6 +75,7 @@ goldenPieces(J), J < 3,
  readCoordinates(X, Y),
  validateCoordinates(CurrentBoard, X, Y, goldenPlayer),
  replaceInBoard(X, Y, goldenPiece, CurrentBoard, NewBoard),
+ asserta( (position(X,Y, goldenPiece)) ),
  G1 is G+1,
  assert(goldenPieces(G1)),
  !.
@@ -84,20 +87,21 @@ goldenPieces(J), J < 3,
  readCoordinates(X, Y),
  validateCoordinates(CurrentBoard, X, Y, goldenPlayer),
  replaceInBoard(X, Y, goldenPiece, CurrentBoard, NewBoard),
+ asserta( (position(X,Y,goldenPiece)) ),
  G1 is G+1,
  assert(goldenPieces(G1)),
  !.
 
 
  placePiece(CurrentBoard, silverPlayer, NewBoard, silverPlayer):-
- silverPieces(J), J =< 3 %aqui e 20
- ,
+ silverPieces(J), J =< 3, %aqui e 20
   retract(silverPieces(G)),
  write('New Piece Coordinates: '),
  repeat,
  readCoordinates(X, Y),
  validateCoordinates(CurrentBoard, X, Y,  silverPlayer),
[…]

### 2015-11-05 game.pl
Minor fixes

@@ -151,7 +151,7 @@ calculateDistances(Xi, Yi, Xf, Yf, DX, DY):-
 	emptySpace(X,Y,X,Yf):- Y> Yf, Y1 is Y-1, emptyCellsBetween(X,Y1,X, Yf).
 	emptySpace(X,Y,X,Yf):- Y < Yf, Y1 is Y+1, emptyCellsBetween(X,Y1,X,Yf).
 
-	emptyCellsBetween(X,Y,X,Y):- findall(Z, position(X,Y,Z), [emptyCell]).
+	emptyCellsBetween(X,Y,X,Y):- findall(Z, position(X,Y,Z), [emptyCell]), !.
 	emptyCellsBetween(X,Y,X,Yf):-Y>Yf,Y1 is Y-1, findall(Z, position(X,Y,Z), [emptyCell]), emptyCellsBetween(X,Y1,X,Yf).
 	emptyCellsBetween(X,Y,X,Yf):-Yf1 is Yf-1, findall(Z, position(X,Yf,Z), [emptyCell]), emptyCellsBetween(X,Y,X,Yf1).
 	emptyCellsBetween(X,Y,Xf,Y):-X>Xf,X1 is X-1, findall(Z, position(X,Y,Z), [emptyCell]), emptyCellsBetween(X1,Y,Xf,Y).

### 2015-11-05 board.pl
misterios da criaçao

@@ -1,115 +1,26 @@
-
-exampleInitial([ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
-                 [' ',' ',' ','p','p','p','p','p',' ',' ',' '],
-                 [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
-                 [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
-                 [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
-                 [' ','p',' ','d',' ','F',' ','d',' ','p',' '],
-                 [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
-                 [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
-                 [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
-                 [' ',' ','p','p','p','p','p','p',' ',' ',' '],
-                 [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '] ]).
-
-exampleMiddle([ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
-                [' ',' ',' ','p','p',' ','d','p',' ',' ',' '],
-                [' ',' ',' ','d',' ',' ',' ',' ',' ',' ',' '],
-                [' ',' ',' ','p','d',' ','d','p',' ',' ',' '],
-                [' ','p',' ',' ',' ','p',' ','d',' ','p',' '],
-                [' ','p',' ','d',' ','F',' ','d',' ','p',' '],
-                [' ','p',' ','d',' ',' ',' ','d',' ','p',' '],
-                [' ','p',' ',' ','d','d','d',' ',' ','p',' '],
-                [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
-                [' ',' ','p','p','p','p','p','p',' ',' ',' '],
-                [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '] ]).
-
-exampleEnd([ [' ',' ',' ',' ',' ',' ',' ',' ',' ',' ',' '],
-                [' ',' ',' ','p','p',' ','d','p',' ',' ',' '],
-                [' ',' ',' ','d',' ',' ',' ',' ',' ',' ',' '],
[…]

### 2015-11-05 game.pl

@@ -16,7 +16,7 @@ piece(flagship).
 player(goldenPlayer).
 player(silverPlayer).
 
-
+currentPlayer(goldenPlayer).
 
 flagships(1).
 
@@ -29,117 +29,65 @@ state(begin).
 state(game).
 state(over).
 
-teste(_):- createBoard(3,3,Board),
-		replaceInBoard(2,2, boda, Board, NewBoard),
-		getCell(NewBoard, 2, 2, boda).
 
-startGame:- assert(goldenPieces(0)), assert(silverPieces(0)),
-			createBoard(11,11,Board),
-			replaceInBoard(5, 5, flagship, Board, NewBoard),
-			setupBoard(NewBoard, NewerBoard),
-			playGame(NewerBoard).
+startGame:- retractall(position(_,_,_)), assert(goldenPieces(0)), assert(silverPieces(0)),
+	setupBoard,
+	playGame.
 
-playGame(Board):-
+playGame:-
 	readPlayer(Player),
-	assert(gameState(Board, Player)),
+	assert(player(Player)),
 	repeat,
-	retract(gameState(CurrentBoard, CurrentPlayer)),
-	takeTurn(Player, CurrentBoard, NewBoard, NewPlayer),
-	printBoard(NewBoard, 11),
-	assert(gameState(NewBoard, Player)),
-	fail. %mudar para winning condition
+	retract(player(NPlayer)),
+	takeTurn(NPlayer, CurrentBoard, NewBoard, NewPlayer),
+	printBoard(0, 0),
+	assert(player(NewPlayer)),
+	fail.
 
 
 
-setupBoard(Board, NewBoard):- assert(gameState(Board, goldenPlayer)),
-					fillBoard(11,11),
-					repeat,
-					retract(gameState(CurrentBoard, CurrentPlayer)),
-					write(CurrentPlayer),
-					placePiece(CurrentBoard, CurrentPlayer, NewBoard, NewPlayer),
-					printBoard(NewBoard, 11),
-					assert(gameState(NewBoard, NewPlayer)),
-					boardFull.
+setupBoard:-
+	fillBoard,
+	asserta(position(5,5,flagship)),
+	placePiece(goldenPlayer, 0).
+
+placePiece(goldenPlayer, 4):- placePiece(silverPlayer, 0).
+placePiece(goldenPlayer, N):- N1 is N+1, N < 4,
[…]

### 2015-11-06 board.pl
Algunos problemas

@@ -3,7 +3,7 @@ symbol(flagship, 'F').
 symbol(silverPiece, 'p').
 symbol(goldenPiece, 'd').
 
-printBoard:- printBoard(0,0).
+printBoard:- printBoard(0,0),!.
 
 printBoard(_,11):- write(' ----------------------------------------------'),nl,
 	write('   0   1   2   3   4   5   6   7   8   9  10'), nl.

### 2015-11-06 game.pl

@@ -8,7 +8,8 @@
 :- dynamic(goldenPieces/1).
 :- dynamic(gameState/2).
 :- dynamic(position/3).
-
+:- dynamic(player/1).
+:- dynamic(currentPlayer/1).
 piece(goldenPiece).
 piece(silverPiece).
 piece(flagship).
@@ -16,35 +17,47 @@ piece(flagship).
 player(goldenPlayer).
 player(silverPlayer).
 
-currentPlayer(goldenPlayer).
-
 flagships(1).
 
 owner(goldenPlayer, goldenPiece).
 owner(silverPlayer, silverPiece).
 owner(goldenPlayer, flagship).
-
+opponent(silverPlayer, goldenPlayer).
+opponent(goldenPlayer, silverPlayer).
 
 state(begin).
 state(game).
 state(over).
 
-
 startGame:- retractall(position(_,_,_)), assert(goldenPieces(0)), assert(silverPieces(0)),
 	setupBoard,
 	playGame.
 
+teste:-fillBoard(0,0), asserta(position(1,1,goldenPiece)), asserta(position(0,0,silverPiece)), asserta(position(0,1,silverPiece)), asserta(position(0,2,silverPiece)), asserta(position(0,3,silverPiece)), asserta(position(4,4,goldenPiece)),
+asserta(position(4,5,goldenPiece)), asserta(position(4,6,goldenPiece)), asserta(position(4,3,goldenPiece)).
+
 playGame:-
 	readPlayer(Player),
-	assert(player(Player)),
+	assert(currentPlayer(Player)),
 	repeat,
-	retract(player(NPlayer)),
-	takeTurn(NPlayer, CurrentBoard, NewBoard, NewPlayer),
-	printBoard(0, 0),
-	assert(player(NewPlayer)),
+	retract(currentPlayer(CurrentPlayer)),
+	write(Player),
+	takeTurn(CurrentPlayer, NewPlayer),
+	assert(currentPlayer(NewPlayer)),
+	printBoard,
 	fail.
 
+takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), doPlayerMovement(goldenPlayer).
[…]

### 2015-11-06 io.pl

@@ -13,4 +13,4 @@ readPlayer(X):-
 	repeat,
 	write('Golden player chooses a player to start the game (silverPlayer. or goldenPlayer.)'), nl,
 	read(X),
-	(X = silverPlayer; X = goldenPlayer).
+	player(X),!.

### 2015-11-06 game.pl
y u derp

@@ -41,14 +41,14 @@ playGame:-
 	assert(currentPlayer(Player)),
 	repeat,
 	retract(currentPlayer(CurrentPlayer)),
-	write(Player),
+	%write(Player),
 	takeTurn(CurrentPlayer, NewPlayer),
 	assert(currentPlayer(NewPlayer)),
 	printBoard,
 	fail.
 
-takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), doPlayerMovement(goldenPlayer).
-takeTurn(silverPlayer, goldenPlayer):-doPlayerMovement(silverPlayer), doPlayerMovement(silverPlayer).
+takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), printBoard, doPlayerMovement(goldenPlayer).
+takeTurn(silverPlayer, goldenPlayer):-doPlayerMovement(silverPlayer), printBoard, doPlayerMovement(silverPlayer).
 
 doPlayerMovement(Player):-write(Player), write(' chooses a piece to move:'), nl,
 repeat,
@@ -115,7 +115,7 @@ calculateDistances(Xi, Yi, Xf, Yf, DX, DY):-
 	emptyCellsBetween(X,Y,Xf,Y):-Xf1 is Xf-1, findall(Z, position(Xf,Y,Z), [emptyCell]), emptyCellsBetween(X,Y,Xf1,Y).
 
 doPlay(X,Y,Xf,Yf,Player):- validMove(X,Y,Xf,Yf,Player),
-	owner(Player,Piece),retract(position(X,Y,Piece)),asserta(position(X,Y,Piece)).
+	owner(Player,Piece),retract(position(X,Y,Piece)),asserta(position(Xf,Yf,Piece)).
 
 doPlay(X,Y,Xf,Yf,Player):- validCapture(X,Y,Xf,Yf,Player),
 	owner(Player,Piece), retract(position(X,Y,Piece)),owner(Player,OpponentPiece), retract(position(Xf,Yf,OpponentPiece)), asserta(position(Xf,Yf,Piece)).

### 2015-11-06 game.pl
Correçao corrigida

@@ -118,7 +118,7 @@ doPlay(X,Y,Xf,Yf,Player):- validMove(X,Y,Xf,Yf,Player),
 	owner(Player,Piece),retract(position(X,Y,Piece)),asserta(position(Xf,Yf,Piece)).
 
 doPlay(X,Y,Xf,Yf,Player):- validCapture(X,Y,Xf,Yf,Player),
-	owner(Player,Piece), retract(position(X,Y,Piece)),owner(Player,OpponentPiece), retract(position(Xf,Yf,OpponentPiece)), asserta(position(Xf,Yf,Piece)).
+	owner(Player,Piece), retract(position(X,Y,Piece)),opponent(Player, Opponent), owner(Opponent,OpponentPiece), retract(position(Xf,Yf,OpponentPiece)), asserta(position(Xf,Yf,Piece)).
 
 validPlay(X,Y,Xf,Yf,Player):- (validMove(X,Y,Xf,Yf,Player);validCapture(X,Y,Xf,Yf,Player)).
 

### 2015-11-07 game.pl
turno acaba quando jogador dourado move flagship

@@ -10,6 +10,7 @@
 :- dynamic(position/3).
 :- dynamic(player/1).
 :- dynamic(currentPlayer/1).
+:- dynamic(moved/1).
 piece(goldenPiece).
 piece(silverPiece).
 piece(flagship).
@@ -34,7 +35,7 @@ startGame:- retractall(position(_,_,_)), assert(goldenPieces(0)), assert(silverP
 	playGame.
 
 teste:-fillBoard(0,0), asserta(position(1,1,goldenPiece)), asserta(position(0,0,silverPiece)), asserta(position(0,1,silverPiece)), asserta(position(0,2,silverPiece)), asserta(position(0,3,silverPiece)), asserta(position(4,4,goldenPiece)),
-asserta(position(4,5,goldenPiece)), asserta(position(4,6,goldenPiece)), asserta(position(4,3,goldenPiece)).
+asserta(position(4,5,goldenPiece)), asserta(position(4,6,goldenPiece)), asserta(position(4,3,goldenPiece)), asserta(position(5,5,flagship)).
 
 playGame:-
 	readPlayer(Player),
@@ -42,12 +43,13 @@ playGame:-
 	repeat,
 	retract(currentPlayer(CurrentPlayer)),
 	%write(Player),
+	retractall(moved(_)),
 	takeTurn(CurrentPlayer, NewPlayer),
 	assert(currentPlayer(NewPlayer)),
 	printBoard,
 	fail.
 
-takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), printBoard, doPlayerMovement(goldenPlayer).
+takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), ( (\+moved(flagship), printBoard, doPlayerMovement(goldenPlayer)); moved(flagship) ).
 takeTurn(silverPlayer, goldenPlayer):-doPlayerMovement(silverPlayer), printBoard, doPlayerMovement(silverPlayer).
 
 doPlayerMovement(Player):-write(Player), write(' chooses a piece to move:'), nl,
@@ -57,6 +59,8 @@ nl,write('destination:'),nl,
 repeat,
 readCoordinates(Xf,Yf),
 validPlay(Xi,Yi,Xf,Yf,Player),!,
+position(Xi,Yi,Piece),
+asserta(moved(Piece)),
 doPlay(Xi,Yi,Xf,Yf,Player).
 
 setupBoard:-

### 2015-11-07 board.pl
did main menu and play menu. got screen clearing command to make game less messy

@@ -3,7 +3,7 @@ symbol(flagship, 'F').
 symbol(silverPiece, 'p').
 symbol(goldenPiece, 'd').
 
-printBoard:- printBoard(0,0),!.
+printBoard:- cls, printBoard(0,0),!.
 
 printBoard(_,11):- write(' ----------------------------------------------'),nl,
 	write('   0   1   2   3   4   5   6   7   8   9  10'), nl.

### 2015-11-07 game.pl

@@ -29,6 +29,10 @@ state(begin).
 state(game).
 state(over).
 
+breakthru:-
+mainMenu.
+
+
 startGame:- retractall(position(_,_,_)), assert(goldenPieces(0)), assert(silverPieces(0)),
 	setupBoard,
 	playGame.
@@ -47,7 +51,7 @@ playGame:-
 	printBoard,
 	fail.
 
-takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), printBoard, doPlayerMovement(goldenPlayer).
+takeTurn(goldenPlayer, silverPlayer):-doPlayerMovement(goldenPlayer), printBoard, doPlayerMovement(goldenPlayer).
 takeTurn(silverPlayer, goldenPlayer):-doPlayerMovement(silverPlayer), printBoard, doPlayerMovement(silverPlayer).
 
 doPlayerMovement(Player):-write(Player), write(' chooses a piece to move:'), nl,

### 2015-11-07 io.pl

@@ -14,3 +14,27 @@ readPlayer(X):-
 	write('Golden player chooses a player to start the game (silverPlayer. or goldenPlayer.)'), nl,
 	read(X),
 	player(X),!.
+
+cls:-write('\e[2J').
+
+mainMenu:-
+write('Welcome to Breakthru!'), nl, nl,
+write('What would you like to do: '), nl,
+write('1. Play game'), nl,
+write('2. Tutorial'), nl,
+write('3. Exit'), nl,
+repeat,
+getInt(Option, 'option'),
+(Option = 1, cls, playMenu;
+Option = 2, cls, tutorial;
+Option = 3).
+
+playMenu:-
+write('How would you like to play: '), nl,
+write('1. Player vs Player'), nl,
+write('2. Player vs Bot'), nl,
+write('3. Bot vs Bot'), nl,
+getInt(Option, 'option'),
+(Option = 1, cls, startGame;
+Option = 2, write('Under construction');
+Option = 3, write('Under construction')).

### 2015-11-07 board.pl
renamed clear command

@@ -3,7 +3,7 @@ symbol(flagship, 'F').
 symbol(silverPiece, 'p').
 symbol(goldenPiece, 'd').
 
-printBoard:- cls, printBoard(0,0),!.
+printBoard:- clearScreen, printBoard(0,0),!.
 
 printBoard(_,11):- write(' ----------------------------------------------'),nl,
 	write('   0   1   2   3   4   5   6   7   8   9  10'), nl.

### 2015-11-07 io.pl

@@ -15,7 +15,7 @@ readPlayer(X):-
 	read(X),
 	player(X),!.
 
-cls:-write('\e[2J').
+clearScreen:-write('\e[2J').
 
 mainMenu:-
 write('Welcome to Breakthru!'), nl, nl,
@@ -25,8 +25,8 @@ write('2. Tutorial'), nl,
 write('3. Exit'), nl,
 repeat,
 getInt(Option, 'option'),
-(Option = 1, cls, playMenu;
-Option = 2, cls, tutorial;
+(Option = 1, clearScreen, playMenu;
+Option = 2, clearScreen, tutorial;
 Option = 3).
 
 playMenu:-
@@ -35,6 +35,6 @@ write('1. Player vs Player'), nl,
 write('2. Player vs Bot'), nl,
 write('3. Bot vs Bot'), nl,
 getInt(Option, 'option'),
-(Option = 1, cls, startGame;
+(Option = 1, clearScreen, startGame;
 Option = 2, write('Under construction');
 Option = 3, write('Under construction')).

### 2015-11-07 game.pl
kind of AI

@@ -45,14 +45,20 @@ playGame:-
 	assert(currentPlayer(Player)),
 	repeat,
 	retract(currentPlayer(CurrentPlayer)),
-	%write(Player),
 	takeTurn(CurrentPlayer, NewPlayer),
 	assert(currentPlayer(NewPlayer)),
 	printBoard,
 	fail.
 
-takeTurn(goldenPlayer, silverPlayer):-doPlayerMovement(goldenPlayer), printBoard, doPlayerMovement(goldenPlayer).
-takeTurn(silverPlayer, goldenPlayer):-doPlayerMovement(silverPlayer), printBoard, doPlayerMovement(silverPlayer).
+takeTurn(goldenPlayer, silverPlayer):-
+(playerGolden(human), doPlayerMovement(goldenPlayer), printBoard, doPlayerMovement(goldenPlayer);
+playerGolden(bot), validPlay(X, Y, Xf, Yf, goldenPlayer), doPlay(X, Y, Xf, Yf, goldenPlayer), printBoard,
+validPlay(X1, Y1, X1f, Y1f, goldenPlayer), doPlay(X1, Y1, X1f, Y1f, goldenPlayer), printBoard).
+
+takeTurn(silverPlayer, goldenPlayer):-
+(playerSilver(human), doPlayerMovement(silverPlayer), printBoard, doPlayerMovement(silverPlayer);
+playerSilver(bot), validPlay(X, Y, Xf, Yf, silverPlayer), doPlay(X, Y, Xf, Yf, silverPlayer), printBoard,
+validPlay(X1, Y1, X1f, Y1f, silverPlayer), doPlay(X1, Y1, X1f, Y1f, silverPlayer), printBoard).
 
 doPlayerMovement(Player):-write(Player), write(' chooses a piece to move:'), nl,
 repeat,

### 2015-11-07 io.pl

@@ -35,6 +35,6 @@ write('1. Player vs Player'), nl,
 write('2. Player vs Bot'), nl,
 write('3. Bot vs Bot'), nl,
 getInt(Option, 'option'),
-(Option = 1, clearScreen, startGame;
-Option = 2, write('Under construction');
-Option = 3, write('Under construction')).
+(Option = 1, clearScreen, assert(playerGolden(human)), assert(playerSilver(human)), startGame;
+Option = 2, clearScreen, assert(playerGolden(human)), assert(playerSilver(bot)), startGame;
+Option = 3, clearScreen, assert(playerGolden(bot)), assert(playerSilver(bot)), startGame).

### 2015-11-07 game.pl
cuts save the day

@@ -53,12 +53,12 @@ playGame:-
 takeTurn(goldenPlayer, silverPlayer):-
 (playerGolden(human), doPlayerMovement(goldenPlayer), printBoard, doPlayerMovement(goldenPlayer);
 playerGolden(bot), validPlay(X, Y, Xf, Yf, goldenPlayer), doPlay(X, Y, Xf, Yf, goldenPlayer), printBoard,
-validPlay(X1, Y1, X1f, Y1f, goldenPlayer), doPlay(X1, Y1, X1f, Y1f, goldenPlayer), printBoard).
+validPlay(X1, Y1, X1f, Y1f, goldenPlayer), doPlay(X1, Y1, X1f, Y1f, goldenPlayer), printBoard), !.
 
 takeTurn(silverPlayer, goldenPlayer):-
 (playerSilver(human), doPlayerMovement(silverPlayer), printBoard, doPlayerMovement(silverPlayer);
 playerSilver(bot), validPlay(X, Y, Xf, Yf, silverPlayer), doPlay(X, Y, Xf, Yf, silverPlayer), printBoard,
-validPlay(X1, Y1, X1f, Y1f, silverPlayer), doPlay(X1, Y1, X1f, Y1f, silverPlayer), printBoard).
+validPlay(X1, Y1, X1f, Y1f, silverPlayer), doPlay(X1, Y1, X1f, Y1f, silverPlayer), printBoard), !.
 
 doPlayerMovement(Player):-write(Player), write(' chooses a piece to move:'), nl,
 repeat,

### 2015-11-07 game.pl
Mergezinho

@@ -11,6 +11,9 @@
 :- dynamic(player/1).
 :- dynamic(currentPlayer/1).
 :- dynamic(moved/1).
+:- dynamic(moved/2).
+:- dynamic(captured/0).
+
 piece(goldenPiece).
 piece(silverPiece).
 piece(flagship).
@@ -42,19 +45,18 @@ playGame:-
 	assert(currentPlayer(Player)),
 	repeat,
 	retract(currentPlayer(CurrentPlayer)),
-	%write(Player),
-	retractall(moved(_)),
+	retractall(moved(_)), retractall(moved(_,_)),retractall(captured),
 	takeTurn(CurrentPlayer, NewPlayer),
 	assert(currentPlayer(NewPlayer)),
 	printBoard,
 	fail.
 
-takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), ( (\+moved(flagship), printBoard, doPlayerMovement(goldenPlayer)); moved(flagship) ).
-takeTurn(silverPlayer, goldenPlayer):-doPlayerMovement(silverPlayer), printBoard, doPlayerMovement(silverPlayer).
+takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), ( (\+moved(flagship), \+captured, printBoard, doPlayerMovement(goldenPlayer)); (moved(flagship);captured) ),!.
+takeTurn(silverPlayer, goldenPlayer):-doPlayerMovement(silverPlayer), printBoard,  ((\+captured, doPlayerMovement(silverPlayer)); (captured) ),!.
 
-doPlayerMovement(Player):-write(Player), write(' chooses a piece to move:'), nl,
-repeat,
-readCoordinates(Xi,Yi),owner(Player,Piece),position(Xi,Yi,Piece), !,
+doPlayerMovement(Player):-
+repeat,write(Player), write(' chooses a piece to move:'), nl,
+readCoordinates(Xi,Yi),owner(Player,Piece),position(Xi,Yi,Piece),\+moved(Xi,Yi), !,
 nl,write('destination:'),nl,
 repeat,
 readCoordinates(Xf,Yf),
@@ -119,23 +121,23 @@ calculateDistances(Xi, Yi, Xf, Yf, DX, DY):-
[…]

### 2015-11-07 game.pl
O jogo acaba

@@ -53,10 +53,11 @@ playGame:-
 	takeTurn(CurrentPlayer, NewPlayer),
 	assert(currentPlayer(NewPlayer)),
 	printBoard,
-	fail.
+	wonGame(Victor),!, clearScreen,
+	write(Victor),write('won the game!').
 
-takeTurn(goldenPlayer,silverPlayer):- doPlayerMovement(goldenPlayer), ( (\+moved(flagship), \+captured, printBoard, doPlayerMovement(goldenPlayer)); (moved(flagship);captured) ),!.
-takeTurn(silverPlayer, goldenPlayer):-doPlayerMovement(silverPlayer), printBoard,  ((\+captured, doPlayerMovement(silverPlayer)); (captured) ),!.
+wonGame(goldenPlayer):- position(_,0,flagship);position(0,_,flagship);position(11,_,flagship);position(_,11,flagship);\+position(_,_,silverPiece).
+wonGame(silverPlayer):- \+position(_,_,flagship).
 
 takeTurn(goldenPlayer, silverPlayer):-
 (playerGolden(human), doPlayerMovement(goldenPlayer), printBoard,( (\+moved(flagship), \+captured, printBoard, doPlayerMovement(goldenPlayer)); (moved(flagship);captured) );

### 2015-11-07 game.pl
random plays 50% more random

@@ -1,4 +1,6 @@
 :- use_module(library(random)).
+:- use_module(library(system)).
+:- use_module(library(lists)).
 :- include('board.pl').
 :- include('util.pl').
 :- include('io.pl').
@@ -34,6 +36,7 @@ mainMenu.
 
 
 startGame:- retractall(position(_,_,_)), assert(goldenPieces(0)), assert(silverPieces(0)),
+	setRandSeed,
 	setupBoard,
 	playGame.
 
@@ -52,13 +55,13 @@ playGame:-
 
 takeTurn(goldenPlayer, silverPlayer):-
 (playerGolden(human), doPlayerMovement(goldenPlayer), printBoard, doPlayerMovement(goldenPlayer);
-playerGolden(bot), validPlay(X, Y, Xf, Yf, goldenPlayer), doPlay(X, Y, Xf, Yf, goldenPlayer), printBoard,
-validPlay(X1, Y1, X1f, Y1f, goldenPlayer), doPlay(X1, Y1, X1f, Y1f, goldenPlayer), printBoard), !.
+playerGolden(bot), randomPlay(goldenPlayer, Pred1), Pred =.. Pred1, Pred, printBoard,
+randomPlay(goldenPlayer, Pred2), PredD =.. Pred2, PredD, doPlay(X1, Y1, X1f, Y1f, goldenPlayer), printBoard), !.
 
 takeTurn(silverPlayer, goldenPlayer):-
 (playerSilver(human), doPlayerMovement(silverPlayer), printBoard, doPlayerMovement(silverPlayer);
-playerSilver(bot), validPlay(X, Y, Xf, Yf, silverPlayer), doPlay(X, Y, Xf, Yf, silverPlayer), printBoard,
-validPlay(X1, Y1, X1f, Y1f, silverPlayer), doPlay(X1, Y1, X1f, Y1f, silverPlayer), printBoard), !.
+playerSilver(bot), randomPlay(silverPlayer, Pred1), Pred =.. Pred1, Pred, printBoard,
+randomPlay(silverPlayer, Pred2), PredD =.. Pred2, PredD, doPlay(X1, Y1, X1f, Y1f, silverPlayer), printBoard), !.
 
 doPlayerMovement(Player):-write(Player), write(' chooses a piece to move:'), nl,
 repeat,
[…]

### 2015-11-07 util.pl

@@ -1,2 +1,4 @@
-coinToss(X):-
-	random(0, 2, X).
+setRandSeed:-
+	now(Time), S is Time mod 30269,
+	getrand(random(X, Y, Z, _)),
+	setrand(random(S, X, Y, Z)), !.

### 2015-11-07 game.pl
corrigido um erro simpes

@@ -148,6 +148,8 @@ doPlay(X,Y,Xf,Yf,Player):- validCapture(X,Y,Xf,Yf,Player),
 validPlay(X,Y,Xf,Yf,Player):- (validMove(X,Y,Xf,Yf,Player);validCapture(X,Y,Xf,Yf,Player)),\+moved(X,Y).
 
 validCapture(X,Y,Xf,Yf,Player):-
+ owner(Player, Piece),
+ position(X,Y,Piece),
  opponent(Player, Opponent),
  owner(Opponent,OpponentPiece),
  position(Xf,Yf,OpponentPiece),

### 2015-11-07 game.pl
fixed AI

@@ -64,13 +64,11 @@ wonGame(silverPlayer):- \+position(_,_,flagship).
 
 takeTurn(goldenPlayer, silverPlayer):-
 (playerGolden(human), doPlayerMovement(goldenPlayer), printBoard,( (\+moved(flagship), \+captured, printBoard, doPlayerMovement(goldenPlayer)); (moved(flagship);captured) );
-playerGolden(bot), validPlay(X, Y, Xf, Yf, goldenPlayer), doPlay(X, Y, Xf, Yf, goldenPlayer), printBoard,
-randomPlay(goldenPlayer, Pred1), Pred =.. Pred1, Pred, (\+moved(flagship),\+captured,randomPlay(goldenPlayer, Pred1), Pred =.. Pred1, Pred);(moved(flagship);captured), printBoard), !.
+playerGolden(bot), randomPlay(goldenPlayer, Pred1), Pred =.. Pred1, Pred, printBoard, (\+moved(flagship),\+captured, randomMove(goldenPlayer, Pred2), PredD =.. Pred2, PredD);(moved(flagship);captured), printBoard), !.
 
 takeTurn(silverPlayer, goldenPlayer):-
 (playerSilver(human), doPlayerMovement(silverPlayer), printBoard, ((\+captured,doPlayerMovement(silverPlayer));captured);
-playerSilver(bot), validPlay(X, Y, Xf, Yf, silverPlayer), doPlay(X, Y, Xf, Yf, silverPlayer), printBoard,
-randomPlay(goldenPlayer, Pred1), Pred =.. Pred1, Pred, (captured;(\+captured,randomPlay(silverPlayer, Pred1), Pred =.. Pred1, Pred)), printBoard), !.
+playerSilver(bot), randomPlay(silverPlayer, Pred1), Pred =.. Pred1, Pred, (captured;(\+captured,randomMove(silverPlayer, Pred2), PredD =.. Pred2, PredD)), printBoard), !.
 
 
 doPlayerMovement(Player):-
@@ -161,8 +159,16 @@ position(Xf,Yf, emptyCell),
 findall(Z, position(Xf,Yf,Z), [emptyCell]),
 emptySpace(X,Y,Xf,Yf).
 
-randomPlay(CurrPlayer, Pred):-
+randomMove(CurrPlayer, Pred):-
[…]

### 2015-11-08 game.pl
bots place dem pieces

@@ -35,7 +35,7 @@ state(begin).
 state(game).
 state(over).
 
-breakthru:-
+breakthru:-retractall(playerGolden(_)), retractall(playerSilver(_)),
 mainMenu.
 
 
@@ -72,15 +72,15 @@ playerSilver(bot), randomPlay(silverPlayer, Pred1), Pred =.. Pred1, Pred, (captu
 
 
 doPlayerMovement(Player):-
-repeat,write(Player), write(' chooses a piece to move:'), nl,
-readCoordinates(Xi,Yi),owner(Player,Piece),position(Xi,Yi,Piece),\+moved(Xi,Yi), !,
-nl,write('destination:'),nl,
-repeat,
-readCoordinates(Xf,Yf),
-validPlay(Xi,Yi,Xf,Yf,Player),!,
-position(Xi,Yi,Piece),
-asserta(moved(Piece)),
-doPlay(Xi,Yi,Xf,Yf,Player).
+	repeat,write(Player), write(' chooses a piece to move:'), nl,
+	readCoordinates(Xi,Yi),owner(Player,Piece),position(Xi,Yi,Piece),\+moved(Xi,Yi), !,
+	nl,write('destination:'),nl,
+	repeat,
+	readCoordinates(Xf,Yf),
+	validPlay(Xi,Yi,Xf,Yf,Player),!,
+	position(Xi,Yi,Piece),
+	asserta(moved(Piece)),
+	doPlay(Xi,Yi,Xf,Yf,Player).
 
 setupBoard:-
 	fillBoard,
@@ -88,14 +88,32 @@ setupBoard:-
 	placePiece(goldenPlayer, 0).
 
 placePiece(goldenPlayer, 4):- placePiece(silverPlayer, 0).
-placePiece(goldenPlayer, N):- N1 is N+1, N < 4,
+placePiece(goldenPlayer, N):-playerGolden(human),
+	N1 is N+1, N < 4,
 	write('New piece for golden player:'), nl,
 	repeat,
 	readCoordinates(X,Y),
 	validateCoordinates(X,Y, goldenPlayer),!,
 	asserta(position(X,Y,goldenPiece)),
-		printBoard,!,
+	printBoard,!,
 	placePiece(goldenPlayer, N1).
+
+placePiece(goldenPlayer,N):-
+	playerGolden(bot),
+	N1 is N+1, N < 4,
[…]

### 2015-11-08 game.pl
detatanço de flagships comunas em fuga

@@ -63,12 +63,17 @@ wonGame(goldenPlayer):- position(_,0,flagship);position(0,_,flagship);position(1
 wonGame(silverPlayer):- \+position(_,_,flagship).
 
 takeTurn(goldenPlayer, silverPlayer):-
-(playerGolden(human), doPlayerMovement(goldenPlayer), printBoard,( (\+moved(flagship), \+captured, printBoard, doPlayerMovement(goldenPlayer)); (moved(flagship);captured) );
-playerGolden(bot), randomPlay(goldenPlayer, Pred1), Pred =.. Pred1, Pred, printBoard, (\+moved(flagship),\+captured, randomMove(goldenPlayer, Pred2), PredD =.. Pred2, PredD);(moved(flagship);captured), printBoard), !.
+playerGolden(human), doPlayerMovement(goldenPlayer), printBoard,( (\+moved(flagship), \+captured, printBoard, doPlayerMovement(goldenPlayer)); (moved(flagship);captured)),!.
+
+takeTurn(goldenPlayer, silverPlayer):-
+playerGolden(bot), difficulty(goldenPlayer, random), randomPlay(goldenPlayer, Pred1), Pred =.. Pred1, Pred, printBoard,
+(\+moved(flagship),\+captured, randomMove(goldenPlayer, Pred2), PredD =.. Pred2, PredD);((moved(flagship);captured), printBoard), !.
 
 takeTurn(silverPlayer, goldenPlayer):-
-(playerSilver(human), doPlayerMovement(silverPlayer), printBoard, ((\+captured,doPlayerMovement(silverPlayer));captured);
-playerSilver(bot), randomPlay(silverPlayer, Pred1), Pred =.. Pred1, Pred, (captured;(\+captured,randomMove(silverPlayer, Pred2), PredD =.. Pred2, PredD)), printBoard), !.
+playerSilver(human), doPlayerMovement(silverPlayer), printBoard, ((\+captured,doPlayerMovement(silverPlayer));captured),!.
+
+takeTurn(silverPlayer, goldenPlayer):-
[…]

### 2015-11-08 io.pl

@@ -34,7 +34,17 @@ write('How would you like to play: '), nl,
 write('1. Player vs Player'), nl,
 write('2. Player vs Bot'), nl,
 write('3. Bot vs Bot'), nl,
+repeat,
 getInt(Option, 'option'),
 (Option = 1, clearScreen, assert(playerGolden(human)), assert(playerSilver(human)), startGame;
-Option = 2, clearScreen, assert(playerGolden(human)), assert(playerSilver(bot)), startGame;
-Option = 3, clearScreen, assert(playerGolden(bot)), assert(playerSilver(bot)), startGame).
+Option = 2, clearScreen, assert(playerGolden(human)), assert(playerSilver(bot)), botDiffMenu(silverPlayer), startGame;
+Option = 3, clearScreen, assert(playerGolden(bot)), botDiffMenu(goldenPlayer), assert(playerSilver(bot)), botDiffMenu(silverPlayer), startGame).
+
+botDiffMenu(Player):-
+	write('Choose bot difficulty for '), write(Player), write(': '), nl,
+	write('1. Random Mode'), nl,
+	write('2. Greedy Mode'), nl,
+	repeat,
+	getInt(Option, 'option'),
+	(Option = 1, clearScreen, assert(difficulty(Player, random));
+	Option = 2, clearScreen, assert(difficulty(Player, greedy))).

### 2015-11-08 game.pl
silverPlayer now with more intelligence

@@ -35,7 +35,7 @@ state(begin).
 state(game).
 state(over).
 
-breakthru:-retractall(playerGolden(_)), retractall(playerSilver(_)),
+breakthru:-retractall(playerGolden(_)), retractall(playerSilver(_)),retractall(difficulty(_,_)),
 mainMenu.
 
 
@@ -73,8 +73,15 @@ takeTurn(silverPlayer, goldenPlayer):-
 playerSilver(human), doPlayerMovement(silverPlayer), printBoard, ((\+captured,doPlayerMovement(silverPlayer));captured),!.
 
 takeTurn(silverPlayer, goldenPlayer):-
-playerSilver(bot), randomPlay(silverPlayer, Pred1), Pred =.. Pred1, Pred, (captured;(\+captured,randomMove(silverPlayer, Pred2), PredD =.. Pred2, PredD)), printBoard, !.
+playerSilver(bot), difficulty(silverPlayer, random), randomPlay(silverPlayer, Pred1), Pred =.. Pred1, Pred, (captured;(\+captured,randomMove(silverPlayer, Pred2), PredD =.. Pred2, PredD)), printBoard, !.
 
+takeTurn(silverPlayer, goldenPlayer):-playerSilver(bot), difficulty(silverPlayer, greedy),
+( ( blockFlagship(X,Y,Xf,Yf), doPlay(X,Y,Xf,Yf,silverPlayer));
+	(validCapture(X,Y,Xf,Yf, silverPlayer), doPlay(X,Y,Xf,Yf,silverPlayer));
+	(randomMove(silverPlayer, Pred1), Pred =.. Pred1, Pred)),!,
+(  captured;
+	 (\+captured,  blockFlagship(X,Y,Xf,Yf), doPlay(X,Y,Xf,Yf,silverPlayer));
+	 (randomMove(silverPlayer, Pred1), Pred =.. Pred1, Pred)).
 
 doPlayerMovement(Player):-
 	repeat,write(Player), write(' chooses a piece to move:'), nl,

### 2015-11-08 game.pl
goldenPlayer tambem já é um burgues ambicioso

@@ -41,7 +41,7 @@ mainMenu.
 
 startGame:- retractall(position(_,_,_)), assert(goldenPieces(0)), assert(silverPieces(0)),
 	setRandSeed,
-	setupBoard,
+	setupBoard,	
 	playGame.
 
 teste:-fillBoard(0,0), asserta(position(1,1,goldenPiece)), asserta(position(0,0,silverPiece)), asserta(position(0,1,silverPiece)), asserta(position(0,2,silverPiece)), asserta(position(0,3,silverPiece)), asserta(position(4,4,goldenPiece)),
@@ -55,11 +55,11 @@ playGame:-
 	retractall(moved(_)), retractall(moved(_,_)),retractall(captured),
 	takeTurn(CurrentPlayer, NewPlayer),
 	assert(currentPlayer(NewPlayer)),
-	printBoard,
+	printBoard, get_char(_),
 	wonGame(Victor),!, clearScreen,
-	write(Victor),write('won the game!').
+	write(Victor),write(' won the game!').
 
-wonGame(goldenPlayer):- position(_,0,flagship);position(0,_,flagship);position(11,_,flagship);position(_,11,flagship);\+position(_,_,silverPiece).
+wonGame(goldenPlayer):- position(_,0,flagship);position(0,_,flagship);position(10,_,flagship);position(_,10,flagship);\+position(_,_,silverPiece).
 wonGame(silverPlayer):- \+position(_,_,flagship).
 
 takeTurn(goldenPlayer, silverPlayer):-
@@ -67,7 +67,16 @@ playerGolden(human), doPlayerMovement(goldenPlayer), printBoard,( (\+moved(flags
 
 takeTurn(goldenPlayer, silverPlayer):-
 playerGolden(bot), difficulty(goldenPlayer, random), randomPlay(goldenPlayer, Pred1), Pred =.. Pred1, Pred, printBoard,
-(\+moved(flagship),\+captured, randomMove(goldenPlayer, Pred2), PredD =.. Pred2, PredD);((moved(flagship);captured), printBoard), !.
+((\+moved(flagship),\+captured, randomMove(goldenPlayer, Pred2), PredD =.. Pred2, PredD);((moved(flagship);captured), printBoard)), !.
+
+takeTurn(goldenPlayer, silverPlayer):-
[…]

### 2015-11-08 game.pl
goldenplayer e broqueado ao inicio

@@ -41,7 +41,7 @@ mainMenu.
 
 startGame:- retractall(position(_,_,_)), assert(goldenPieces(0)), assert(silverPieces(0)),
 	setRandSeed,
-	setupBoard,	
+	setupBoard,
 	playGame.
 
 teste:-fillBoard(0,0), asserta(position(1,1,goldenPiece)), asserta(position(0,0,silverPiece)), asserta(position(0,1,silverPiece)), asserta(position(0,2,silverPiece)), asserta(position(0,3,silverPiece)), asserta(position(4,4,goldenPiece)),
@@ -55,7 +55,7 @@ playGame:-
 	retractall(moved(_)), retractall(moved(_,_)),retractall(captured),
 	takeTurn(CurrentPlayer, NewPlayer),
 	assert(currentPlayer(NewPlayer)),
-	printBoard, get_char(_),
+	printBoard,
 	wonGame(Victor),!, clearScreen,
 	write(Victor),write(' won the game!').
 
@@ -108,7 +108,7 @@ setupBoard:-
 	asserta(position(5,5,flagship)),
 	placePiece(goldenPlayer, 0).
 
-placePiece(goldenPlayer, 12):- placePiece(silverPlayer, 0).
+placePiece(goldenPlayer, 3):- placePiece(silverPlayer, 0).
 placePiece(goldenPlayer, N):-playerGolden(human),
 	N1 is N+1, N < 12,
 	write('New piece for golden player:'), nl,
@@ -121,11 +121,13 @@ placePiece(goldenPlayer, N):-playerGolden(human),
 
 placePiece(goldenPlayer,N):-
 	playerGolden(bot),
-	N1 is N+1, N < 12,
+	N1 is N+1, N < 3,
 	randomPlacement(goldenPlayer, X, Y),
 	asserta(position(X,Y,goldenPiece)),
 	printBoard,!,
 	placePiece(goldenPlayer, N1).
+
+	placePiece(silverPlayer,N):-playerSilver(bot), difficulty(silverPlayer, greedy),blockFlagshipPlacement, N1 is N-1, printBoard, !, placePiece(silverPlayer, N-1).
 
 	placePiece(silverPlayer,N):-
 		playerSilver(bot),
@@ -238,3 +240,11 @@ flagshipCanEscape(X,Y):-
[…]

### 2015-11-08 game.pl
mais cenes bonites

@@ -48,7 +48,7 @@ teste:-fillBoard(0,0), asserta(position(1,1,goldenPiece)), asserta(position(0,0,
 asserta(position(4,5,goldenPiece)), asserta(position(4,6,goldenPiece)), asserta(position(4,3,goldenPiece)), asserta(position(5,5,flagship)).
 
 playGame:-
-	readPlayer(Player),
+	readPlayer(Player, 'goldenPlayer chooses a player to start the game (goldenPlayer. or silverPlayer.).'),
 	assert(currentPlayer(Player)),
 	repeat,
 	retract(currentPlayer(CurrentPlayer)),
@@ -108,7 +108,7 @@ setupBoard:-
 	asserta(position(5,5,flagship)),
 	placePiece(goldenPlayer, 0).
 
-placePiece(goldenPlayer, 3):- placePiece(silverPlayer, 0).
+placePiece(goldenPlayer, 12):- placePiece(silverPlayer, 0).
 placePiece(goldenPlayer, N):-playerGolden(human),
 	N1 is N+1, N < 12,
 	write('New piece for golden player:'), nl,
@@ -121,7 +121,7 @@ placePiece(goldenPlayer, N):-playerGolden(human),
 
 placePiece(goldenPlayer,N):-
 	playerGolden(bot),
-	N1 is N+1, N < 3,
+	N1 is N+1, N < 12,
 	randomPlacement(goldenPlayer, X, Y),
 	asserta(position(X,Y,goldenPiece)),
 	printBoard,!,

### 2015-11-08 io.pl

@@ -9,9 +9,9 @@ readCoordinates(X1, Y1):-
  getInt(X1, 'x'),nl,
  getInt(Y1, 'y').
 
-readPlayer(X):-
+readPlayer(X, Prompt):-
 	repeat,
-	write('Golden player chooses a player to start the game (silverPlayer. or goldenPlayer.)'), nl,
+	write(Prompt), nl,
 	read(X),
 	player(X),!.
 
@@ -37,7 +37,7 @@ write('3. Bot vs Bot'), nl,
 repeat,
 getInt(Option, 'option'),
 (Option = 1, clearScreen, assert(playerGolden(human)), assert(playerSilver(human)), startGame;
-Option = 2, clearScreen, assert(playerGolden(human)), assert(playerSilver(bot)), botDiffMenu(silverPlayer), startGame;
+Option = 2, clearScreen, chooseSideMenu, startGame;
 Option = 3, clearScreen, assert(playerGolden(bot)), botDiffMenu(goldenPlayer), assert(playerSilver(bot)), botDiffMenu(silverPlayer), startGame).
 
 botDiffMenu(Player):-
@@ -48,3 +48,8 @@ botDiffMenu(Player):-
 	getInt(Option, 'option'),
 	(Option = 1, clearScreen, assert(difficulty(Player, random));
 	Option = 2, clearScreen, assert(difficulty(Player, greedy))).
+
+chooseSideMenu:-
+	readPlayer(Player, 'Choose which side you want to be on (silverPlayer. or goldenPlayer.)'),
+	(Player = goldenPlayer, assert(playerGolden(human)), assert(playerSilver(bot)), botDiffMenu(silverPlayer);
+	Player = silverPlayer, assert(playerSilver(human)), assert(playerGolden(bot)), botDiffMenu(goldenPlayer)).

### 2015-11-08 game.pl
fully AI'd

@@ -5,7 +5,7 @@
 :- include('util.pl').
 :- include('io.pl').
 
-:- dynamic(flagships/1).
+
 :- dynamic(silverPieces/1).
 :- dynamic(goldenPieces/1).
 :- dynamic(gameState/2).
@@ -48,8 +48,9 @@ teste:-fillBoard(0,0), asserta(position(1,1,goldenPiece)), asserta(position(0,0,
 asserta(position(4,5,goldenPiece)), asserta(position(4,6,goldenPiece)), asserta(position(4,3,goldenPiece)), asserta(position(5,5,flagship)).
 
 playGame:-
-	readPlayer(Player, 'goldenPlayer chooses a player to start the game (goldenPlayer. or silverPlayer.).'),
-	assert(currentPlayer(Player)),
+	( (\+playerGolden(bot),readPlayer(Player, 'goldenPlayer chooses a player to start the game (goldenPlayer. or silverPlayer.).')); (playerGolden(bot), random(0, 2, N), nth0(N, [goldenPlayer, silverPlayer], Player)
+	)),
+assert(currentPlayer(Player)),
 	repeat,
 	retract(currentPlayer(CurrentPlayer)),
 	retractall(moved(_)), retractall(moved(_,_)),retractall(captured),

### 2015-11-08 doc/template.tex
relatorio final começado e concluido mas nam emmeiado

@@
 \title{\Huge\textbf{Breakthru}\linebreak\linebreak\linebreak
-\Large\textbf{Relatório Intercalar}\linebreak\linebreak
+\Large\textbf{Relatório Final}\linebreak\linebreak
 \linebreak\linebreak
@@
 
-%************************************************************************************************
-%************************************************************************************************
+\section{Resumo}
+No contexto da unidade curricular Programação em Lógica do MIEIC foi solicitada a implementação de um jogo de tabuleiro a ser selecionado pelo grupo.
 
+Para a implementação deste jogo o grupo mobilizou os seus conhecimentos acerca de PROLOG e das capacidades generativas que o mesmo proporciona e que se revelaram de elevada importância para permitir o desenvolvimento de um jogo do ponto de vista declarativo, tornando-se apenas necessário explicitar em PROLOG as regras de funcionamento do jogo.
+
+O acima exposto evidencia a acentuada diferença paradigmática da programação nesta linguagem, que representou para o grupo um desafio mas também uma oportunidade de adquirir destreza mental no que à programação em lógica diz respeito.
+
+\tableofcontents
 \newpage
 
-%Todas as figuras devem ser referidas no texto. %\ref{fig:codigoFigura}
-%
-%%Exemplo de código para inserção de figuras
-%%\begin{figure}[h!]
-%%\begin{center}
-%%escolher entre uma das seguintes três linhas:
-%%\includegraphics[height=20cm,width=15cm]{path relativo da imagem}
-%%\includegraphics[scale=0.5]{path relativo da imagem}
-%%\includegraphics{path relativo da imagem}
-%%\caption{legenda da figura}
-%%\label{fig:codigoFigura}
-%%\end{center}
-%%\end{figure}
-%
-%
-%\textit{Para escrever em itálico}
-%\textbf{Para escrever em negrito}
-%Para escrever em letra normal
-%``Para escrever texto entre aspas''
-%
-%Para fazer parágrafo, deixar uma linha em branco.
-%
-%Como fazer bullet points:
-%\begin{itemize}
-	%\item Item1
-	%\item Item2
-%\end{itemize}
-%
-%Como enumerar itens:
-%\begin{enumerate}
-	%\item Item 1
-	%\item Item 2
-%\end{enumerate}
-%
-%\begin{quote}``Isto é uma citação''\end{quote}
+\section{Introdução}
+Foi proposta a implementação do jogo \textit{Breakthru} em PROLOG com uma interface textual e com a possibilidade de ser jogado em três modos diversos: humano contra humano, humano contra máquina e máquina contra máquina.
+Foi solicitado que o mecanismo de jogo da máquina apresentasse dois níveis distintos de dificuldade.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%
+O presente relatório principia por apresentar o jogo desenvolvido, o seu contexto histórico e as regras que o regem.
+De seguida são apresentadas as principais estruturações programáticas do mesmo, do ponto de vista do modelo de dados usado e dos predicados lógicos fundamentais da implementação desenvolvida.
+Em particular é apresentada a representação interna do estado do jogo, os predicados usados para apresentar externamente essa mesma representação, bem como diversos predicados e restante lógica de movimentação de peças que permita verificar o cumprimento das regras por parte dos jogadores humanos e a escolha de jogadas por parte do computador, quando em controlo de um dos jogadores.
+São depois tecidas considerações sobre a interface com o utilizador e suas funcionalidades e implementação.
+
+Encerra-se o relatório com uma breve reflexão sobre o trabalho desenvolvido, suas limitações e aspetos que ficam em aberto para eventuais futuras melhorias.
+
 \section{O Jogo Breakthru}
@@
 \subsubsection{Movimentos e Capturas}
-Cada jogador deve, na sua vez, efetuar dois movimentos ou uma captura.
+Cada jogador deve, na sua vez, efetuar dois movimentos ou uma captura. O jogador dourado escolhe quem deve jogar primeiro.
 
@@
 
+\section{Conclusão}
+Ficaram, no entender do grupo, fortemente evidenciadas as vantagens situacionais do uso do paradigma em causa para abordar a classe de problemas que frequentemente surgem no desenvolvimento de simuladores de jogos de tabuleiros.
+Com efeito, o trabalho desenvolvido evidencia que, recorrendo a capacidades que são exclusivas da programação em lógica, se torna possível exprimir de forma sucinta funcionalidades de um programa que se revelariam de elevada complexidade, do ponto de vista da implementação.
+O PROLOG permite em larga medida que o programador prescinda da necessidade de especificar um modelo de execução, o que constitui, regra geral, o maior desafio para a implementação deste tipo de programas.
+
+O trabalho apresenta como principal limitação o facto de a inteligência artificial implementada para o segundo nível de dificuldade não apresentar uma função de avaliação do tabuleiro plenamente desenvolvida.
+A lógica de avaliação implementada é essencialmente binária: a máquina procura a cada passo garantir somente que não vai perder no passo seguinte e que captura ao adversário o máximo de peças possíveis, sendo que os restantes movimentos são selecionados, à falta de heurística desenvolvida, de uma forma aleatória.
+Considerou-se o desenvolvimento de tal função de avaliação e do seu uso num algoritmo \textit{Minimax}, mas limitações ao nível do tempo e da compreensão plena deste algoritmo, que não foi extensivamente abordado nas aulas da Unidade Curricular, não no-lo permitiram.
+
 \end{document}

### 2015-11-08 game.pl
Moved some stuff around

@@ -4,6 +4,7 @@
 :- include('board.pl').
 :- include('util.pl').
 :- include('io.pl').
+:- include('bot.pl').
 
 
 :- dynamic(silverPieces/1).
@@ -205,47 +206,9 @@ position(Xf,Yf, emptyCell),
 findall(Z, position(Xf,Yf,Z), [emptyCell]),
 emptySpace(X,Y,Xf,Yf).
 
-randomMove(CurrPlayer, Pred):-
-findall([Xi,Yi,Xf,Yf], validMove(Xi,Yi,Xf,Yf,CurrPlayer), Possiveis),
-length(Possiveis, N),
-random(0, N, NList),
-nth0(NList, Possiveis, ChosenPlay),
-append(ChosenPlay, [CurrPlayer], AlmostPred),
-append([doPlay], AlmostPred, Pred).
-
-randomPlay(CurrPlayer, Pred):-
-findall([Xi,Yi,Xf,Yf], validPlay(Xi,Yi,Xf,Yf,CurrPlayer), Possiveis),
-length(Possiveis, N),
-random(0, N, NList),
-nth0(NList, Possiveis, ChosenPlay),
-append(ChosenPlay, [CurrPlayer], AlmostPred),
-append([doPlay], AlmostPred, Pred).
-
-randomPlacement(CurrPlayer, X,Y):-
-	findall([Xp, Yp], possiblePlace(CurrPlayer, Xp, Yp), Possiveis),
-	length(Possiveis, N),
-	random(0, N, NList),
-	nth0(NList, Possiveis, [X,Y]).
-
 flagshipCanEscape(X,Y):-
 	position(Fx,Fy,flagship),
 	( ( emptySpace(Fx,Fy,X,10), Y is 10);
 		(emptySpace(Fx,Fy,X,0), Y is 0);
 		(emptySpace(Fx,Fy,10,Y), X is 10);
 		(emptySpace(Fx,Fy,0,Y), X is 0) ).
-
-	blockFlagship(X,Y,Xj,Yj):-
-		position(Xf,Yf,flagship),
-		flagshipCanEscape(Xe,Ye),
-		((Xe = Xf, Ye > Yf, validMove(X,Y,Xe, Yj,silverPlayer), Yj > Yf, Xj is Xe ); %escapa para baixo
-		(Xe = Xf, Ye < Yf, validMove(X,Y,Xe, Yj,silverPlayer), Yj < Yf, Xj is Xe ); %escapa para cima
[…]

### 2015-11-23 proj1/game.pl
algumas consideraçoes sobre o segundo projeto a realizar na disciplina em causa que é a de Programação Em Logica, onde se ensina fundamentalmente o ploG

@@ -105,5 +105,5 @@ validateCoordinates(Board, X, Y, silverPlayer):-
 	
 	
 takeTurn(Player):-
-	write("Choose the piece you want to move: "), nl,
+	write('Choose the piece you want to move: '), nl,
 	getCoordinates(X, Y, Board),

### 2015-11-23 proj2/algumasconsideracoessobreoproblemaemestudo.pl

@@ -1,0 +1,14 @@
+:-use_module(library(clpfd)).
+
+%entradas
+	% professor - (Nome-[Lista de linguas |_]-[Lista de dias-rendimentoporhora|_])
+	% candidatura - (NomeAluno-Lingua)
+	% linguas - (Lingua-Preço)
+
+% saidas
+	% ([Lingua-dia-professor-carga horaria|TAIL])
+	% lucro maninho
+	
+	
+escola_de_linguas():-
+	

### 2015-11-25 bot.pl
remove the bot.pl file that was in excess

@@ -1,37 +1,0 @@
-randomMove(CurrPlayer, Pred):-
-findall([Xi,Yi,Xf,Yf], validMove(Xi,Yi,Xf,Yf,CurrPlayer), Possiveis),
-length(Possiveis, N),
-random(0, N, NList),
-nth0(NList, Possiveis, ChosenPlay),
-append(ChosenPlay, [CurrPlayer], AlmostPred),
-append([doPlay], AlmostPred, Pred).
-
-randomPlay(CurrPlayer, Pred):-
-findall([Xi,Yi,Xf,Yf], validPlay(Xi,Yi,Xf,Yf,CurrPlayer), Possiveis),
-length(Possiveis, N),
-random(0, N, NList),
-nth0(NList, Possiveis, ChosenPlay),
-append(ChosenPlay, [CurrPlayer], AlmostPred),
-append([doPlay], AlmostPred, Pred).
-
-randomPlacement(CurrPlayer, X,Y):-
-	findall([Xp, Yp], possiblePlace(CurrPlayer, Xp, Yp), Possiveis),
-	length(Possiveis, N),
-	random(0, N, NList),
-	nth0(NList, Possiveis, [X,Y]).
-
-blockFlagship(X,Y,Xj,Yj):-
-  position(Xf,Yf,flagship),
-  flagshipCanEscape(Xe,Ye),
-  ((Xe = Xf, Ye > Yf, validMove(X,Y,Xe, Yj,silverPlayer), Yj > Yf, Xj is Xe ); %escapa para baixo
-  (Xe = Xf, Ye < Yf, validMove(X,Y,Xe, Yj,silverPlayer), Yj < Yf, Xj is Xe ); %escapa para cima
-  (Ye = Yf, Xe > Xf, validMove(X,Y,Xj, Ye,silverPlayer), Xj > Xf, Yj is Ye ); %escapa para a direita
-  (Ye = Yf, Xe < Xf, validMove(X,Y,Xj, Ye,silverPlayer), Xj < Xf, Yj is Ye )). %escapa para a esquerda
-
-blockFlagshipPlacement:-
-  position(Xf,Yf,flagship),
-  flagshipCanEscape(Xe,Ye),
-  ((Xe = Xf, Ye > Yf, asserta(position(Xe,10,silverPiece))); %escapa para baixo
-  (Xe = Xf, Ye < Yf, asserta(position(Xe,0,silverPiece))); %escapa para cima
[…]

### 2015-12-08 proj2/algumasconsideracoessobreoproblemaemestudo.pl
algumas restriçoes

@@ -2,28 +2,87 @@
 
 %entradas
 	% professor - (indiceP-[Lista de linguas-rendimentoporhora|_]-[Lista de dias|_])
-	% candidatura - (indiceC-indiceL)
+	% candidatura - (Lingua-Numero)
 	% linguas - (indiceL-Preço)
 
 % saidas
 	% ([lingua|TAIL])
-	% lucro 
+	% lucro
 %restriçao
 	% 1 - max 15 horas/semana/professor
 	% 2 - minimo 2 horas por semana/linguas
 	% 3 - maximo 4 horas/semana/lingua
 	% 4 - 15 alunos/lingua
 	% 5 - em cada dia nao pode haver mais do que x aulas em cada intervalo inicio-fim, para todos os intervalos
-	
+
 clearScreen:-write('\e[2J').
-	
+
+
+professoresQueDaoLingua(Lingua, [], []).
+professoresQueDaoLingua(Lingua,  [ [Prof, Linguas,_] | OutrosProfs], Lista):-
+	member(Lingua-_, Linguas),
+	professoresQueDaoLingua(Lingua, OutrosProfs, Lista2),
+	append([Prof], Lista2, Lista).
+
+professoresQueDaoLingua(Lingua,  [ [Prof, Linguas,_] | OutrosProfs], Lista):-
+	professoresQueDaoLingua(Lingua, OutrosProfs, Lista).
+
+
+testeProfs(Sol):-
+	professoresQueDaoLingua(5, [[0, [0-1,1-2],_], [1, [0-2],_], [2, [3-2],_]], Sol).
+
+
+
+
+initialize_days_list([],_,_,_).
+initialize_days_list([H|T], N, NProfessores, FdSetDeProfs):-
+	H = N-Hora-Prof,
+	Prof in_set FdSetDeProfs,
+	domain([Hora], 0, 4),
+	N1 is N+1,
+	initialize_days_list(T, N1, NProfessores, FdSetDeProfs).
+
+
+initialize_solution([], _,_,_).
+initialize_solution([H|T], Lingua, NProfessores, Professores):-
+	H = Lingua-Lista,
+	length(Lista, 5),
+	professoresQueDaoLingua(Lingua, Professores, ListaDeProfs),
+	list_to_fdset([-1|ListaDeProfs], FdSetDeProfs),
+	initialize_days_list(Lista, 0, NProfessores,FdSetDeProfs),
[…]

### 2015-12-08 proj2/listaexemplo.pl

@@ -1,0 +1,3 @@
+[[0, [0-19, 1-19, 2-10],[0,1,2,3,4,5,6]], [1, [4-25],[0,1,2]]]
+
+[0-1,1-1,1-2,1-3]

### 2015-12-08 proj2/algumasconsideracoessobreoproblemaemestudo.pl
restriçao 15 horas semanais/professor

@@ -36,7 +36,8 @@ testeProfs(Sol):-
 
 initialize_days_list([],_,_,_).
 initialize_days_list([H|T], N, NProfessores, FdSetDeProfs):-
-	H = N-Hora-Prof,
+	H = [N,Hora,Prof],
+
 	Prof in_set FdSetDeProfs,
 	domain([Hora], 0, 4),
 	N1 is N+1,
@@ -68,8 +69,43 @@ excludeEmpty([Lingua-Horarios|Solucoes], Candidaturas):-
 	excludeEmpty([Lingua-Horarios|Solucoes], Candidaturas):-
 		excludeEmpty(Solucoes, Candidaturas).
 
+
+% [ 0-[ [0, 2, 4], [1, 3, 5] ], 1-[[0,4,5], [1,3,5],[2,6,6] ]  ]
+
 testeExclude:-
 	excludeEmpty([0-[]|T], [0-0]).
+labelAll([]).
+labelAll([H|T]):-
+	H = _-[Dia1,Dia2,Dia3,Dia4,Dia5],
+	labeling([], Dia1),
+	labeling([], Dia2),
+	labeling([], Dia3),
+	labeling([], Dia4),
+	labeling([], Dia5),
+	labelAll(T).
+
+restrict_teacher_hours(Solucao, []).
+restrict_teacher_hours(Solucao, [Professor|T]):-
+	Professor=[Indice,_,_],
+	restrict_teacher(Solucao, Indice, 0),
+	restrict_teacher_hours(Solucao, T).
+
+%
+restrict_teacher([], _, Acum):- Acum #< 15, write(Acum).
+restrict_teacher([H|T], Indice, Acum):-
+	H = IndiceL-Dias,
+	teacher_times(Dias, Indice, Time),
+	Acum1 #= Time+Acum,
+	restrict_teacher(T, Indice, Acum1).
+
+
+teacher_times([], Indice, 0).
+teacher_times([[IndiceDia, Horas, ProfessorResponsavel]|Dias], Indice, Time):-
+	ProfessorResponsavel #= Indice #<=> B,
+	TempoQueOProfDaEstaAula #= B*Horas,
+	teacher_times(Dias, Indice, TempoRestante),
+	Time #= TempoQueOProfDaEstaAula+TempoRestante.
+
 
 
 
@@ -79,10 +115,45 @@ escola_de_linguas(Professores, Candidaturas, Linguas, Lucro, Solucao):-
 	length(Linguas, NLinguas),
[…]

### 2015-12-09 proj2/doc/llncs.tex
started article. as far as I can tell we need miktex for this.

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}

\tableofcontents
%
\mainmatter              % start of the contributions
%
\title{Solving an optimization problem employing restraint programming}
%
\titlerunning{Restraint Programming}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{João Cabral\inst{1} \and João Mota\inst{2}}
%
\authorrunning{João Cabral \and João Mota} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{João Cabral, João Mota}
%
\institute{Faculdade de Engenharia da Universidade do Porto, Porto, Rua Roberto Frias s/n 4200-465, Portugal,\\
\email{(up201303462, up201304395)@fe.up.pt}}

\maketitle              % typeset the title of the contribution

\begin{abstract}
This paper goes over the theoretical and practical aspects of the resolution of an optimization problem,
namely the synthesis of a schedule for a language school, employing restraint programming, more precisely, 
SICSTUS Prolog's restraint programming capabilities.
The paper contains a description of the problem, a detailed analysis of the aproach used to solve it, 
an explanation of how the solution should be interpreted, a description of obtained results and, finally, conclusions.

\keywords{optimization, restraint programming, PROLOG}
\end{abstract}
%
\section{Fixed-Period Problems: The Sublinear Case}
%
With this chapter, the preliminaries are over, and we begin the search
for periodic solutions to Hamiltonian systems. All this will be done in
the convex case; that is, we shall study the boundary-value problem
\begin{eqnarray*}
  \dot{x}&=&JH' (t,x)\\
  x(0) &=& x(T)
\end{eqnarray*}
with $H(t,\cdot)$ a convex function of $x$, going to $+\infty$ when
$\left\|x\right\| \to \infty$.

%
\subsection{Autonomous Systems}
%
In this section, we will consider the case when the Hamiltonian $H(x)$
is autonomous. For the sake of simplicity, we shall also assume that it
is $C^{1}$.

We shall first consider the question of nontriviality, within the
general framework of
$\left(A_{\infty},B_{\infty}\right)$-subquadratic Hamiltonians. In
the second subsection, we shall look into the special case when $H$ is
$\left(0,b_{\infty}\right)$-subquadratic,
and we shall try to derive additional information.
%
\subsubsection{The General Case: Nontriviality.}
%
We assume that $H$ is
$\left(A_{\infty},B_{\infty}\right)$-sub\-qua\-dra\-tic at infinity,
for some constant symmetric matrices $A_{\infty}$ and $B_{\infty}$,
with $B_{\infty}-A_{\infty}$ positive definite. Set:
\begin{eqnarray}
\gamma :&=&{\rm smallest\ eigenvalue\ of}\ \ B_{\infty} - A_{\infty} \\
  \lambda : &=& {\rm largest\ negative\ eigenvalue\ of}\ \
  J \frac{d}{dt} +A_{\infty}\ .
\end{eqnarray}

Theorem~\ref{ghou:pre} tells us that if $\lambda +\gamma < 0$, the
boundary-value problem:
\begin{equation}
\begin{array}{rcl}
  \dot{x}&=&JH' (x)\\
  x(0)&=&x (T)
\end{array}
\end{equation}
has at least one solution
$\overline{x}$, which is found by minimizing the dual
action functional:
\begin{equation}
  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt
\end{equation}
on the range of $\Lambda$, which is a subspace $R (\Lambda)_{L}^{2}$
with finite codimension. Here
\begin{equation}
  N(x) := H(x) - \frac{1}{2} \left(A_{\infty} x,x\right)
\end{equation}
is a convex function, and
\begin{equation}
  N(x) \le \frac{1}{2}
  \left(\left(B_{\infty} - A_{\infty}\right) x,x\right)
  + c\ \ \ \forall x\ .
\end{equation}

%
\begin{proposition}
Assume $H'(0)=0$ and $ H(0)=0$. Set:
\begin{equation}
  \delta := \liminf_{x\to 0} 2 N (x) \left\|x\right\|^{-2}\ .
  \label{eq:one}
\end{equation}

If $\gamma < - \lambda < \delta$,
the solution $\overline{u}$ is non-zero:
\begin{equation}
  \overline{x} (t) \ne 0\ \ \ \forall t\ .
\end{equation}
\end{proposition}
%
\begin{proof}
Condition (\ref{eq:one}) means that, for every
$\delta ' > \delta$, there is some $\varepsilon > 0$ such that
\begin{equation}
  \left\|x\right\| \le \varepsilon \Rightarrow N (x) \le
  \frac{\delta '}{2} \left\|x\right\|^{2}\ .
\end{equation}

It is an exercise in convex analysis, into which we shall not go, to
show that this implies that there is an $\eta > 0$ such that
\begin{equation}
  f\left\|x\right\| \le \eta
  \Rightarrow N^{\ast} (y) \le \frac{1}{2\delta '}
  \left\|y\right\|^{2}\ .
  \label{eq:two}
\end{equation}

\begin{figure}
\vspace{2.5cm}
\caption{This is the caption of the figure displaying a white eagle and
a white horse on a snow field}
\end{figure}

Since $u_{1}$ is a smooth function, we will have
$\left\|hu_{1}\right\|_\infty \le \eta$
for $h$ small enough, and inequality (\ref{eq:two}) will hold,
yielding thereby:
\begin{equation}
  \psi (hu_{1}) \le \frac{h^{2}}{2}
  \frac{1}{\lambda} \left\|u_{1} \right\|_{2}^{2} + \frac{h^{2}}{2}
  \frac{1}{\delta '} \left\|u_{1}\right\|^{2}\ .
\end{equation}

If we choose $\delta '$ close enough to $\delta$, the quantity
$\left(\frac{1}{\lambda} + \frac{1}{\delta '}\right)$
will be negative, and we end up with
\begin{equation}
  \psi (hu_{1}) < 0\ \ \ \ \ {\rm for}\ \ h\ne 0\ \ {\rm small}\ .
\end{equation}

On the other hand, we check directly that $\psi (0) = 0$. This shows
that 0 cannot be a minimizer of $\psi$, not even a local one.
So $\overline{u} \ne 0$ and
$\overline{u} \ne \Lambda_{o}^{-1} (0) = 0$. \qed
\end{proof}
%
\begin{corollary}
Assume $H$ is $C^{2}$ and
$\left(a_{\infty},b_{\infty}\right)$-subquadratic at infinity. Let
$\xi_{1},\allowbreak\dots,\allowbreak\xi_{N}$  be the
equilibria, that is, the solutions of $H' (\xi ) = 0$.
Denote by $\omega_{k}$
the smallest eigenvalue of $H'' \left(\xi_{k}\right)$, and set:
\begin{equation}
  \omega : = {\rm Min\,} \left\{\omega_{1},\dots,\omega_{k}\right\}\ .
\end{equation}
If:
\begin{equation}
  \frac{T}{2\pi} b_{\infty} <
  - E \left[- \frac{T}{2\pi}a_{\infty}\right] <
  \frac{T}{2\pi}\omega
  \label{eq:three}
\end{equation}
then minimization of $\psi$ yields a non-constant $T$-periodic solution
$\overline{x}$.
\end{corollary}
%

We recall once more that by the integer part $E [\alpha ]$ of
$\alpha \in \bbbr$, we mean the $a\in \bbbz$
such that $a< \alpha \le a+1$. For instance,
if we take $a_{\infty} = 0$, Corollary 2 tells
us that $\overline{x}$ exists and is
non-constant provided that:

\begin{equation}
  \frac{T}{2\pi} b_{\infty} < 1 < \frac{T}{2\pi}
\end{equation}
or
\begin{equation}
  T\in \left(\frac{2\pi}{\omega},\frac{2\pi}{b_{\infty}}\right)\ .
  \label{eq:four}
\end{equation}

%
\begin{proof}
The spectrum of $\Lambda$ is $\frac{2\pi}{T} \bbbz +a_{\infty}$. The
largest negative eigenvalue $\lambda$ is given by
$\frac{2\pi}{T}k_{o} +a_{\infty}$,
where
\begin{equation}
  \frac{2\pi}{T}k_{o} + a_{\infty} < 0
  \le \frac{2\pi}{T} (k_{o} +1) + a_{\infty}\ .
\end{equation}
Hence:
\begin{equation}
  k_{o} = E \left[- \frac{T}{2\pi} a_{\infty}\right] \ .
\end{equation}

The condition $\gamma < -\lambda < \delta$ now becomes:
\begin{equation}
  b_{\infty} - a_{\infty} <
  - \frac{2\pi}{T} k_{o} -a_{\infty} < \omega -a_{\infty}
\end{equation}
which is precisely condition (\ref{eq:three}).\qed
\end{proof}
%

\begin{lemma}
Assume that $H$ is $C^{2}$ on $\bbbr^{2n} \setminus \{ 0\}$ and
that $H'' (x)$ is non-de\-gen\-er\-ate for any $x\ne 0$. Then any local
minimizer $\widetilde{x}$ of $\psi$ has minimal period $T$.
\end{lemma}
%
\begin{proof}
We know that $\widetilde{x}$, or
$\widetilde{x} + \xi$ for some constant $\xi
\in \bbbr^{2n}$, is a $T$-periodic solution of the Hamiltonian system:
\begin{equation}
  \dot{x} = JH' (x)\ .
\end{equation}

There is no loss of generality in taking $\xi = 0$. So
$\psi (x) \ge \psi (\widetilde{x} )$
for all $\widetilde{x}$ in some neighbourhood of $x$ in
$W^{1,2} \left(\bbbr / T\bbbz ; \bbbr^{2n}\right)$.

But this index is precisely the index
$i_{T} (\widetilde{x} )$ of the $T$-periodic
solution $\widetilde{x}$ over the interval
$(0,T)$, as defined in Sect.~2.6. So
\begin{equation}
  i_{T} (\widetilde{x} ) = 0\ .
  \label{eq:five}
\end{equation}

Now if $\widetilde{x}$ has a lower period, $T/k$ say,
we would have, by Corollary 31:
\begin{equation}
  i_{T} (\widetilde{x} ) =
  i_{kT/k}(\widetilde{x} ) \ge
  ki_{T/k} (\widetilde{x} ) + k-1 \ge k-1 \ge 1\ .
\end{equation}

This would contradict (\ref{eq:five}), and thus cannot happen.\qed
\end{proof}
%
\paragraph{Notes and Comments.}
The results in this section are a
refined version of \cite{clar:eke};
the minimality result of Proposition
14 was the first of its kind.

To understand the nontriviality conditions, such as the one in formula
(\ref{eq:four}), one may think of a one-parameter family
$x_{T}$, $T\in \left(2\pi\omega^{-1}, 2\pi b_{\infty}^{-1}\right)$
of periodic solutions, $x_{T} (0) = x_{T} (T)$,
with $x_{T}$ going away to infinity when $T\to 2\pi \omega^{-1}$,
which is the period of the linearized system at 0.

\begin{table}
\caption{This is the example table taken out of {\it The
\TeX{}book,} p.\,246}
\begin{center}
\begin{tabular}{r@{\quad}rl}
\hline
\multicolumn{1}{l}{\rule{0pt}{12pt}
                   Year}&\multicolumn{2}{l}{World population}\\[2pt]
\hline\rule{0pt}{12pt}
8000 B.C.  &     5,000,000& \\
  50 A.D.  &   200,000,000& \\
1650 A.D.  &   500,000,000& \\
1945 A.D.  & 2,300,000,000& \\
1980 A.D.  & 4,400,000,000& \\[2pt]
\hline
\end{tabular}
\end{center}
\end{table}
%
\begin{theorem} [Ghoussoub-Preiss]\label{ghou:pre}
Assume $H(t,x)$ is
$(0,\varepsilon )$-subquadratic at
infinity for all $\varepsilon > 0$, and $T$-periodic in $t$
\begin{equation}
  H (t,\cdot )\ \ \ \ \ {\rm is\ convex}\ \ \forall t
\end{equation}
\begin{equation}
  H (\cdot ,x)\ \ \ \ \ {\rm is}\ \ T{\rm -periodic}\ \ \forall x
\end{equation}
\begin{equation}
  H (t,x)\ge n\left(\left\|x\right\|\right)\ \ \ \ \
  {\rm with}\ \ n (s)s^{-1}\to \infty\ \ {\rm as}\ \ s\to \infty
\end{equation}
\begin{equation}
  \forall \varepsilon > 0\ ,\ \ \ \exists c\ :\
  H(t,x) \le \frac{\varepsilon}{2}\left\|x\right\|^{2} + c\ .
\end{equation}

Assume also that $H$ is $C^{2}$, and $H'' (t,x)$ is positive definite
everywhere. Then there is a sequence $x_{k}$, $k\in \bbbn$, of
$kT$-periodic solutions of the system
\begin{equation}
  \dot{x} = JH' (t,x)
\end{equation}
such that, for every $k\in \bbbn$, there is some $p_{o}\in\bbbn$ with:
\begin{equation}
  p\ge p_{o}\Rightarrow x_{pk} \ne x_{k}\ .
\end{equation}
\qed
\end{theorem}
%
\begin{example} [{{\rm External forcing}}]
Consider the system:
\begin{equation}
  \dot{x} = JH' (x) + f(t)
\end{equation}
where the Hamiltonian $H$ is
$\left(0,b_{\infty}\right)$-subquadratic, and the
forcing term is a distribution on the circle:
\begin{equation}
  f = \frac{d}{dt} F + f_{o}\ \ \ \ \
  {\rm with}\ \ F\in L^{2} \left(\bbbr / T\bbbz; \bbbr^{2n}\right)\ ,
\end{equation}
where $f_{o} : = T^{-1}\int_{o}^{T} f (t) dt$. For instance,
\begin{equation}
  f (t) = \sum_{k\in \bbbn} \delta_{k} \xi\ ,
\end{equation}
where $\delta_{k}$ is the Dirac mass at $t= k$ and
$\xi \in \bbbr^{2n}$ is a
constant, fits the prescription. This means that the system
$\dot{x} = JH' (x)$ is being excited by a
series of identical shocks at interval $T$.
\end{example}
%
\begin{definition}
Let $A_{\infty} (t)$ and $B_{\infty} (t)$ be symmetric
operators in $\bbbr^{2n}$, depending continuously on
$t\in [0,T]$, such that
$A_{\infty} (t) \le B_{\infty} (t)$ for all $t$.

A Borelian function
$H: [0,T]\times \bbbr^{2n} \to \bbbr$
is called
$\left(A_{\infty} ,B_{\infty}\right)$-{\it subquadratic at infinity}
if there exists a function $N(t,x)$ such that:
\begin{equation}
  H (t,x) = \frac{1}{2} \left(A_{\infty} (t) x,x\right) + N(t,x)
\end{equation}
\begin{equation}
  \forall t\ ,\ \ \ N(t,x)\ \ \ \ \
  {\rm is\ convex\ with\  respect\  to}\ \ x
\end{equation}
\begin{equation}
  N(t,x) \ge n\left(\left\|x\right\|\right)\ \ \ \ \
  {\rm with}\ \ n(s)s^{-1}\to +\infty\ \ {\rm as}\ \ s\to +\infty
\end{equation}
\begin{equation}
  \exists c\in \bbbr\ :\ \ \ H (t,x) \le
  \frac{1}{2} \left(B_{\infty} (t) x,x\right) + c\ \ \ \forall x\ .
\end{equation}

If $A_{\infty} (t) = a_{\infty} I$ and
$B_{\infty} (t) = b_{\infty} I$, with
$a_{\infty} \le b_{\infty} \in \bbbr$,
we shall say that $H$ is
$\left(a_{\infty},b_{\infty}\right)$-subquadratic
at infinity. As an example, the function
$\left\|x\right\|^{\alpha}$, with
$1\le \alpha < 2$, is $(0,\varepsilon )$-subquadratic at infinity
for every $\varepsilon > 0$. Similarly, the Hamiltonian
\begin{equation}
H (t,x) = \frac{1}{2} k \left\|k\right\|^{2} +\left\|x\right\|^{\alpha}
\end{equation}
is $(k,k+\varepsilon )$-subquadratic for every $\varepsilon > 0$.
Note that, if $k<0$, it is not convex.
\end{definition}
%

\paragraph{Notes and Comments.}
The first results on subharmonics were
obtained by Rabinowitz in \cite{rab}, who showed the existence of
infinitely many subharmonics both in the subquadratic and superquadratic
case, with suitable growth conditions on $H'$. Again the duality
approach enabled Clarke and Ekeland in \cite{clar:eke:2} to treat the
same problem in the convex-subquadratic case, with growth conditions on
$H$ only.

Recently, Michalek and Tarantello (see \cite{mich:tar} and \cite{tar})
have obtained lower bound on the number of subharmonics of period $kT$,
based on symmetry considerations and on pinching estimates, as in
Sect.~5.2 of this article.

%
% ---- Bibliography ----
%
\begin{thebibliography}{5}
%
\bibitem {clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and
boundary-value problems for Hamiltonian systems.
Arch. Rat. Mech. Anal. 78, 315--333 (1982)

\bibitem {clar:eke:2}
Clarke, F., Ekeland, I.:
Solutions p\'{e}riodiques, du
p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
Note CRAS Paris 287, 1013--1015 (1978)

\bibitem {mich:tar}
Michalek, R., Tarantello, G.:
Subharmonic solutions with prescribed minimal
period for nonautonomous Hamiltonian systems.
J. Diff. Eq. 72, 28--55 (1988)

\bibitem {tar}
Tarantello, G.:
Subharmonic solutions for Hamiltonian
systems via a $\bbbz_{p}$ pseudoindex theory.
Annali di Matematica Pura (to appear)

\bibitem {rab}
Rabinowitz, P.:
On subharmonic solutions of a Hamiltonian system.
Comm. Pure Appl. Math. 33, 609--633 (1980)

\end{thebibliography}

%
% second contribution with nearly identical text,
% slightly changed contribution head (all entries
% appear as defaults), and modified bibliography
%
\title{Hamiltonian Mechanics2}

\author{Ivar Ekeland\inst{1} \and Roger Temam\inst{2}}

\institute{Princeton University, Princeton NJ 08544, USA
\and
Universit\'{e} de Paris-Sud,
Laboratoire d'Analyse Num\'{e}rique, B\^{a}timent 425,\\
F-91405 Orsay Cedex, France}

\maketitle
%
% Modify the bibliography environment to call for the author-year
% system. This is done normally with the citeauthoryear option
% for a particular contribution.
\makeatletter
\renewenvironment{thebibliography}[1]
     {\section*{\refname}
      \small
      \list{}%
           {\settowidth\labelwidth{}%
            \leftmargin\parindent
            \itemindent=-\parindent
            \labelsep=\z@
            \if@openbib
              \advance\leftmargin\bibindent
              \itemindent -\bibindent
              \listparindent \itemindent
              \parsep \z@
            \fi
            \usecounter{enumiv}%
            \let\p@enumiv\@empty
            \renewcommand\theenumiv{}}%
      \if@openbib
        \renewcommand\newblock{\par}%
      \else
        \renewcommand\newblock{\hskip .11em \@plus.33em \@minus.07em}%
      \fi
      \sloppy\clubpenalty4000\widowpenalty4000%
      \sfcode`\.=\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
      \def\@cite#1{#1}%
      \def\@lbibitem[#1]#2{\item[]\if@filesw
        {\def\protect##1{\string ##1\space}\immediate
      \write\@auxout{\string\bibcite{#2}{#1}}}\fi\ignorespaces}
\makeatother
%
\begin{abstract}
The abstract should summarize the contents of the paper
using at least 70 and at most 150 words. It will be set in 9-point
font size and be inset 1.0 cm from the right and left margins.
There will be two blank lines before and after the Abstract. \dots
\keywords{graph transformations, convex geometry, lattice computations,
convex polygons, triangulations, discrete geometry}
\end{abstract}
%
\section{Fixed-Period Problems: The Sublinear Case}
%
With this chapter, the preliminaries are over, and we begin the search
for periodic solutions to Hamiltonian systems. All this will be done in
the convex case; that is, we shall study the boundary-value problem
\begin{eqnarray*}
  \dot{x}&=&JH' (t,x)\\
  x(0) &=& x(T)
\end{eqnarray*}
with $H(t,\cdot)$ a convex function of $x$, going to $+\infty$ when
$\left\|x\right\| \to \infty$.

%
\subsection{Autonomous Systems}
%
In this section, we will consider the case when the Hamiltonian $H(x)$
is autonomous. For the sake of simplicity, we shall also assume that it
is $C^{1}$.

We shall first consider the question of nontriviality, within the
general framework of
$\left(A_{\infty},B_{\infty}\right)$-subquadratic Hamiltonians. In
the second subsection, we shall look into the special case when $H$ is
$\left(0,b_{\infty}\right)$-subquadratic,
and we shall try to derive additional information.
%
\subsubsection{The General Case: Nontriviality.}
%
We assume that $H$ is
$\left(A_{\infty},B_{\infty}\right)$-sub\-qua\-dra\-tic at infinity,
for some constant symmetric matrices $A_{\infty}$ and $B_{\infty}$,
with $B_{\infty}-A_{\infty}$ positive definite. Set:
\begin{eqnarray}
\gamma :&=&{\rm smallest\ eigenvalue\ of}\ \ B_{\infty} - A_{\infty} \\
  \lambda : &=& {\rm largest\ negative\ eigenvalue\ of}\ \
  J \frac{d}{dt} +A_{\infty}\ .
\end{eqnarray}

Theorem 21 tells us that if $\lambda +\gamma < 0$, the boundary-value
problem:
\begin{equation}
\begin{array}{rcl}
  \dot{x}&=&JH' (x)\\
  x(0)&=&x (T)
\end{array}
\end{equation}
has at least one solution
$\overline{x}$, which is found by minimizing the dual
action functional:
\begin{equation}
  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt
\end{equation}
on the range of $\Lambda$, which is a subspace $R (\Lambda)_{L}^{2}$
with finite codimension. Here
\begin{equation}
  N(x) := H(x) - \frac{1}{2} \left(A_{\infty} x,x\right)
\end{equation}
is a convex function, and
\begin{equation}
  N(x) \le \frac{1}{2}
  \left(\left(B_{\infty} - A_{\infty}\right) x,x\right)
  + c\ \ \ \forall x\ .
\end{equation}

%
\begin{proposition}
Assume $H'(0)=0$ and $ H(0)=0$. Set:
\begin{equation}
  \delta := \liminf_{x\to 0} 2 N (x) \left\|x\right\|^{-2}\ .
  \label{2eq:one}
\end{equation}

If $\gamma < - \lambda < \delta$,
the solution $\overline{u}$ is non-zero:
\begin{equation}
  \overline{x} (t) \ne 0\ \ \ \forall t\ .
\end{equation}
\end{proposition}
%
\begin{proof}
Condition (\ref{2eq:one}) means that, for every
$\delta ' > \delta$, there is some $\varepsilon > 0$ such that
\begin{equation}
  \left\|x\right\| \le \varepsilon \Rightarrow N (x) \le
  \frac{\delta '}{2} \left\|x\right\|^{2}\ .
\end{equation}

It is an exercise in convex analysis, into which we shall not go, to
show that this implies that there is an $\eta > 0$ such that
\begin{equation}
  f\left\|x\right\| \le \eta
  \Rightarrow N^{\ast} (y) \le \frac{1}{2\delta '}
  \left\|y\right\|^{2}\ .
  \label{2eq:two}
\end{equation}

\begin{figure}
\vspace{2.5cm}
\caption{This is the caption of the figure displaying a white eagle and
a white horse on a snow field}
\end{figure}

Since $u_{1}$ is a smooth function, we will have
$\left\|hu_{1}\right\|_\infty \le \eta$
for $h$ small enough, and inequality (\ref{2eq:two}) will hold,
yielding thereby:
\begin{equation}
  \psi (hu_{1}) \le \frac{h^{2}}{2}
  \frac{1}{\lambda} \left\|u_{1} \right\|_{2}^{2} + \frac{h^{2}}{2}
  \frac{1}{\delta '} \left\|u_{1}\right\|^{2}\ .
\end{equation}

If we choose $\delta '$ close enough to $\delta$, the quantity
$\left(\frac{1}{\lambda} + \frac{1}{\delta '}\right)$
will be negative, and we end up with
\begin{equation}
  \psi (hu_{1}) < 0\ \ \ \ \ {\rm for}\ \ h\ne 0\ \ {\rm small}\ .
\end{equation}

On the other hand, we check directly that $\psi (0) = 0$. This shows
that 0 cannot be a minimizer of $\psi$, not even a local one.
So $\overline{u} \ne 0$ and
$\overline{u} \ne \Lambda_{o}^{-1} (0) = 0$. \qed
\end{proof}
%
\begin{corollary}
Assume $H$ is $C^{2}$ and
$\left(a_{\infty},b_{\infty}\right)$-subquadratic at infinity. Let
$\xi_{1},\allowbreak\dots,\allowbreak\xi_{N}$  be the
equilibria, that is, the solutions of $H' (\xi ) = 0$.
Denote by $\omega_{k}$
the smallest eigenvalue of $H'' \left(\xi_{k}\right)$, and set:
\begin{equation}
  \omega : = {\rm Min\,} \left\{\omega_{1},\dots,\omega_{k}\right\}\ .
\end{equation}
If:
\begin{equation}
  \frac{T}{2\pi} b_{\infty} <
  - E \left[- \frac{T}{2\pi}a_{\infty}\right] <
  \frac{T}{2\pi}\omega
  \label{2eq:three}
\end{equation}
then minimization of $\psi$ yields a non-constant $T$-periodic solution
$\overline{x}$.
\end{corollary}
%

We recall once more that by the integer part $E [\alpha ]$ of
$\alpha \in \bbbr$, we mean the $a\in \bbbz$
such that $a< \alpha \le a+1$. For instance,
if we take $a_{\infty} = 0$, Corollary 2 tells
us that $\overline{x}$ exists and is
non-constant provided that:

\begin{equation}
  \frac{T}{2\pi} b_{\infty} < 1 < \frac{T}{2\pi}
\end{equation}
or
\begin{equation}
  T\in \left(\frac{2\pi}{\omega},\frac{2\pi}{b_{\infty}}\right)\ .
  \label{2eq:four}
\end{equation}

%
\begin{proof}
The spectrum of $\Lambda$ is $\frac{2\pi}{T} \bbbz +a_{\infty}$. The
largest negative eigenvalue $\lambda$ is given by
$\frac{2\pi}{T}k_{o} +a_{\infty}$,
where
\begin{equation}
  \frac{2\pi}{T}k_{o} + a_{\infty} < 0
  \le \frac{2\pi}{T} (k_{o} +1) + a_{\infty}\ .
\end{equation}
Hence:
\begin{equation}
  k_{o} = E \left[- \frac{T}{2\pi} a_{\infty}\right] \ .
\end{equation}

The condition $\gamma < -\lambda < \delta$ now becomes:
\begin{equation}
  b_{\infty} - a_{\infty} <
  - \frac{2\pi}{T} k_{o} -a_{\infty} < \omega -a_{\infty}
\end{equation}
which is precisely condition (\ref{2eq:three}).\qed
\end{proof}
%

\begin{lemma}
Assume that $H$ is $C^{2}$ on $\bbbr^{2n} \setminus \{ 0\}$ and
that $H'' (x)$ is non-de\-gen\-er\-ate for any $x\ne 0$. Then any local
minimizer $\widetilde{x}$ of $\psi$ has minimal period $T$.
\end{lemma}
%
\begin{proof}
We know that $\widetilde{x}$, or
$\widetilde{x} + \xi$ for some constant $\xi
\in \bbbr^{2n}$, is a $T$-periodic solution of the Hamiltonian system:
\begin{equation}
  \dot{x} = JH' (x)\ .
\end{equation}

There is no loss of generality in taking $\xi = 0$. So
$\psi (x) \ge \psi (\widetilde{x} )$
for all $\widetilde{x}$ in some neighbourhood of $x$ in
$W^{1,2} \left(\bbbr / T\bbbz ; \bbbr^{2n}\right)$.

But this index is precisely the index
$i_{T} (\widetilde{x} )$ of the $T$-periodic
solution $\widetilde{x}$ over the interval
$(0,T)$, as defined in Sect.~2.6. So
\begin{equation}
  i_{T} (\widetilde{x} ) = 0\ .
  \label{2eq:five}
\end{equation}

Now if $\widetilde{x}$ has a lower period, $T/k$ say,
we would have, by Corollary 31:
\begin{equation}
  i_{T} (\widetilde{x} ) =
  i_{kT/k}(\widetilde{x} ) \ge
  ki_{T/k} (\widetilde{x} ) + k-1 \ge k-1 \ge 1\ .
\end{equation}

This would contradict (\ref{2eq:five}), and thus cannot happen.\qed
\end{proof}
%
\paragraph{Notes and Comments.}
The results in this section are a
refined version of \cite{2clar:eke};
the minimality result of Proposition
14 was the first of its kind.

To understand the nontriviality conditions, such as the one in formula
(\ref{2eq:four}), one may think of a one-parameter family
$x_{T}$, $T\in \left(2\pi\omega^{-1}, 2\pi b_{\infty}^{-1}\right)$
of periodic solutions, $x_{T} (0) = x_{T} (T)$,
with $x_{T}$ going away to infinity when $T\to 2\pi \omega^{-1}$,
which is the period of the linearized system at 0.

\begin{table}
\caption{This is the example table taken out of {\it The
\TeX{}book,} p.\,246}
\begin{center}
\begin{tabular}{r@{\quad}rl}
\hline
\multicolumn{1}{l}{\rule{0pt}{12pt}
                   Year}&\multicolumn{2}{l}{World population}\\[2pt]
\hline\rule{0pt}{12pt}
8000 B.C.  &     5,000,000& \\
  50 A.D.  &   200,000,000& \\
1650 A.D.  &   500,000,000& \\
1945 A.D.  & 2,300,000,000& \\
1980 A.D.  & 4,400,000,000& \\[2pt]
\hline
\end{tabular}
\end{center}
\end{table}
%
\begin{theorem} [Ghoussoub-Preiss]
Assume $H(t,x)$ is
$(0,\varepsilon )$-subquadratic at
infinity for all $\varepsilon > 0$, and $T$-periodic in $t$
\begin{equation}
  H (t,\cdot )\ \ \ \ \ {\rm is\ convex}\ \ \forall t
\end{equation}
\begin{equation}
  H (\cdot ,x)\ \ \ \ \ {\rm is}\ \ T{\rm -periodic}\ \ \forall x
\end{equation}
\begin{equation}
  H (t,x)\ge n\left(\left\|x\right\|\right)\ \ \ \ \
  {\rm with}\ \ n (s)s^{-1}\to \infty\ \ {\rm as}\ \ s\to \infty
\end{equation}
\begin{equation}
  \forall \varepsilon > 0\ ,\ \ \ \exists c\ :\
  H(t,x) \le \frac{\varepsilon}{2}\left\|x\right\|^{2} + c\ .
\end{equation}

Assume also that $H$ is $C^{2}$, and $H'' (t,x)$ is positive definite
everywhere. Then there is a sequence $x_{k}$, $k\in \bbbn$, of
$kT$-periodic solutions of the system
\begin{equation}
  \dot{x} = JH' (t,x)
\end{equation}
such that, for every $k\in \bbbn$, there is some $p_{o}\in\bbbn$ with:
\begin{equation}
  p\ge p_{o}\Rightarrow x_{pk} \ne x_{k}\ .
\end{equation}
\qed
\end{theorem}
%
\begin{example} [{{\rm External forcing}}]
Consider the system:
\begin{equation}
  \dot{x} = JH' (x) + f(t)
\end{equation}
where the Hamiltonian $H$ is
$\left(0,b_{\infty}\right)$-subquadratic, and the
forcing term is a distribution on the circle:
\begin{equation}
  f = \frac{d}{dt} F + f_{o}\ \ \ \ \
  {\rm with}\ \ F\in L^{2} \left(\bbbr / T\bbbz; \bbbr^{2n}\right)\ ,
\end{equation}
where $f_{o} : = T^{-1}\int_{o}^{T} f (t) dt$. For instance,
\begin{equation}
  f (t) = \sum_{k\in \bbbn} \delta_{k} \xi\ ,
\end{equation}
where $\delta_{k}$ is the Dirac mass at $t= k$ and
$\xi \in \bbbr^{2n}$ is a
constant, fits the prescription. This means that the system
$\dot{x} = JH' (x)$ is being excited by a
series of identical shocks at interval $T$.
\end{example}
%
\begin{definition}
Let $A_{\infty} (t)$ and $B_{\infty} (t)$ be symmetric
operators in $\bbbr^{2n}$, depending continuously on
$t\in [0,T]$, such that
$A_{\infty} (t) \le B_{\infty} (t)$ for all $t$.

A Borelian function
$H: [0,T]\times \bbbr^{2n} \to \bbbr$
is called
$\left(A_{\infty} ,B_{\infty}\right)$-{\it subquadratic at infinity}
if there exists a function $N(t,x)$ such that:
\begin{equation}
  H (t,x) = \frac{1}{2} \left(A_{\infty} (t) x,x\right) + N(t,x)
\end{equation}
\begin{equation}
  \forall t\ ,\ \ \ N(t,x)\ \ \ \ \
  {\rm is\ convex\ with\  respect\  to}\ \ x
\end{equation}
\begin{equation}
  N(t,x) \ge n\left(\left\|x\right\|\right)\ \ \ \ \
  {\rm with}\ \ n(s)s^{-1}\to +\infty\ \ {\rm as}\ \ s\to +\infty
\end{equation}
\begin{equation}
  \exists c\in \bbbr\ :\ \ \ H (t,x) \le
  \frac{1}{2} \left(B_{\infty} (t) x,x\right) + c\ \ \ \forall x\ .
\end{equation}

If $A_{\infty} (t) = a_{\infty} I$ and
$B_{\infty} (t) = b_{\infty} I$, with
$a_{\infty} \le b_{\infty} \in \bbbr$,
we shall say that $H$ is
$\left(a_{\infty},b_{\infty}\right)$-subquadratic
at infinity. As an example, the function
$\left\|x\right\|^{\alpha}$, with
$1\le \alpha < 2$, is $(0,\varepsilon )$-subquadratic at infinity
for every $\varepsilon > 0$. Similarly, the Hamiltonian
\begin{equation}
H (t,x) = \frac{1}{2} k \left\|k\right\|^{2} +\left\|x\right\|^{\alpha}
\end{equation}
is $(k,k+\varepsilon )$-subquadratic for every $\varepsilon > 0$.
Note that, if $k<0$, it is not convex.
\end{definition}
%

\paragraph{Notes and Comments.}
The first results on subharmonics were
obtained by Rabinowitz in \cite{2rab}, who showed the existence of
infinitely many subharmonics both in the subquadratic and superquadratic
case, with suitable growth conditions on $H'$. Again the duality
approach enabled Clarke and Ekeland in \cite{2clar:eke:2} to treat the
same problem in the convex-subquadratic case, with growth conditions on
$H$ only.

Recently, Michalek and Tarantello (see Michalek, R., Tarantello, G.
\cite{2mich:tar} and Tarantello, G. \cite{2tar}) have obtained lower
bound on the number of subharmonics of period $kT$, based on symmetry
considerations and on pinching estimates, as in Sect.~5.2 of this
article.

%
% ---- Bibliography ----
%
\begin{thebibliography}{}
%
\bibitem[1980]{2clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and
boundary-value problems for Hamiltonian systems.
Arch. Rat. Mech. Anal. 78, 315--333 (1982)

\bibitem[1981]{2clar:eke:2}
Clarke, F., Ekeland, I.:
Solutions p\'{e}riodiques, du
p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
Note CRAS Paris 287, 1013--1015 (1978)

\bibitem[1982]{2mich:tar}
Michalek, R., Tarantello, G.:
Subharmonic solutions with prescribed minimal
period for nonautonomous Hamiltonian systems.
J. Diff. Eq. 72, 28--55 (1988)

\bibitem[1983]{2tar}
Tarantello, G.:
Subharmonic solutions for Hamiltonian
systems via a $\bbbz_{p}$ pseudoindex theory.
Annali di Matematica Pura (to appear)

\bibitem[1985]{2rab}
Rabinowitz, P.:
On subharmonic solutions of a Hamiltonian system.
Comm. Pure Appl. Math. 33, 609--633 (1980)

\end{thebibliography}
\clearpage
\addtocmark[2]{Author Index} % additional numbered TOC entry
\renewcommand{\indexname}{Author Index}
\printindex
\clearpage
\addtocmark[2]{Subject Index} % additional numbered TOC entry
\markboth{Subject Index}{Subject Index}
\renewcommand{\indexname}{Subject Index}
\input{subjidx.ind}
\end{document}

### 2015-12-09 proj2/algumasconsideracoessobreoproblemaemestudo.pl
Tentativa de otimizaçao

@@ -18,13 +18,13 @@
 clearScreen:-write('\e[2J').
 
 
-professoresQueDaoLingua(Lingua, [], []).
+professoresQueDaoLingua(_, [], []).
 professoresQueDaoLingua(Lingua,  [ [Prof, Linguas,_] | OutrosProfs], Lista):-
 	member(Lingua-_, Linguas),
 	professoresQueDaoLingua(Lingua, OutrosProfs, Lista2),
 	append([Prof], Lista2, Lista).
 
-professoresQueDaoLingua(Lingua,  [ [Prof, Linguas,_] | OutrosProfs], Lista):-
+professoresQueDaoLingua(Lingua,  [ [_, _,_] | OutrosProfs], Lista):-
 	professoresQueDaoLingua(Lingua, OutrosProfs, Lista).
 
 
@@ -55,7 +55,7 @@ initialize_solution([H|T], Lingua, NProfessores, Professores):-
 	initialize_solution(T, Lingua1, NProfessores, Professores).
 
 forceEmpty([]).
-forceEmpty([Dia-Horas-Prof|T]):-
+forceEmpty([_-Horas-Prof|T]):-
 	Horas in {0},
 	Prof in {-1},
 	forceEmpty(T).
@@ -66,25 +66,23 @@ excludeEmpty([Lingua-Horarios|Solucoes], Candidaturas):-
 	forceEmpty(Horarios),
 	excludeEmpty(Solucoes, Candidaturas).
 
-	excludeEmpty([Lingua-Horarios|Solucoes], Candidaturas):-
+	excludeEmpty([_-_|Solucoes], Candidaturas):-
 		excludeEmpty(Solucoes, Candidaturas).
 
 
 % [ 0-[ [0, 2, 4], [1, 3, 5] ], 1-[[0,4,5], [1,3,5],[2,6,6] ]  ]
 
-testeExclude:-
-	excludeEmpty([0-[]|T], [0-0]).
-labelAll([]).
-labelAll([H|T]):-
+labelAll([],_).
+labelAll([H|T], Cost):-
 	H = _-[Dia1,Dia2,Dia3,Dia4,Dia5],
 	labeling([], Dia1),
 	labeling([], Dia2),
 	labeling([], Dia3),
 	labeling([], Dia4),
-	labeling([], Dia5),
+	labeling([maximize(Cost)], Dia5),
 	labelAll(T).
 
-restrict_teacher_hours(Solucao, []).
[…]

### 2015-12-09 proj2/algumasconsideracoessobreoproblemaemestudo.pl
I think I found it

@@ -1,4 +1,7 @@
 :-use_module(library(clpfd)).
+:-use_module(library(aggregate)).
+
+
 
 %entradas
 	% professor - (indiceP-[Lista de linguas-rendimentoporhora|_]-[Lista de dias|_])
@@ -73,14 +76,17 @@ excludeEmpty([Lingua-Horarios|Solucoes], Candidaturas):-
 % [ 0-[ [0, 2, 4], [1, 3, 5] ], 1-[[0,4,5], [1,3,5],[2,6,6] ]  ]
 
 labelAll([],_).
-labelAll([H|T], Cost):-
-	H = _-[Dia1,Dia2,Dia3,Dia4,Dia5],
-	labeling([], Dia1),
-	labeling([], Dia2),
-	labeling([], Dia3),
-	labeling([], Dia4),
-	labeling([maximize(Cost)], Dia5),
-	labelAll(T).
+labelAll(Sol, Cost):-
+	labelAll(Sol, Cost, []).
+
+labelAll([], Cost, Flat):- labeling(maximize(Cost), Flat).
+labelAll([Indice-[Dia1,Dia2,Dia3,Dia4,Dia5]|Tail], Cost, Flat):-
+	append(Flat, Dia1, Flat1),
+	append(Flat1, Dia2, Flat2),
+	append(Flat2, Dia3, Flat3),
+	append(Flat3, Dia4, Flat4),
+	append(Flat4, Dia5, Flat5),
+	labelAll(Tail, Cost, Flat5).
 
 restrict_teacher_hours(_, []).
 restrict_teacher_hours(Solucao, [Professor|T]):-
@@ -118,8 +124,12 @@ escola_de_linguas(Professores, Candidaturas, Linguas, Lucro, Solucao):-
 	labelAll(Solucao, Lucro).
 
 teste(Solucao, Lucro):-
-%	escola_de_linguas([[0, [0-19, 1-19, 2-10],[0,1,2,3,4,5,6]], [1, [4-25],[0,1,2]]], [0-15, 1-10, 2-3, 3-10,4-1], [0-1,1-1,2-2,3-3,4-4], _, Solucao).
-	escola_de_linguas([[0, [0-10],[0,1]]], [0-10], [0-20], Lucro, Solucao).
[…]

### 2015-12-09 proj2/algumasconsideracoessobreoproblemaemestudo.pl
Problema ilustrado

@@ -79,8 +79,9 @@ labelAll([],_).
 labelAll(Sol, Cost):-
 	labelAll(Sol, Cost, []).
 
-labelAll([], Cost, Flat):- labeling(maximize(Cost), Flat).
+labelAll([], Cost, Flat):-write('gona label'),nl, maximize(labeling([], Flat), Cost).
 labelAll([Indice-[Dia1,Dia2,Dia3,Dia4,Dia5]|Tail], Cost, Flat):-
+	write('labeling'),nl,
 	append(Flat, Dia1, Flat1),
 	append(Flat1, Dia2, Flat2),
 	append(Flat2, Dia3, Flat3),
@@ -120,12 +121,12 @@ escola_de_linguas(Professores, Candidaturas, Linguas, Lucro, Solucao):-
 	initialize_solution(Solucao, 0, NProfessores, Professores),
 	restrict_teacher_hours(Solucao, Professores),
 	make_profit(Solucao, Linguas, Professores, Candidaturas, Lucro),
-	write(Lucro),
-	labelAll(Solucao, Lucro).
+	write(Lucro).
+	%labelAll(Solucao, Lucro).
 
 teste(Solucao, Lucro):-
-%escola_de_linguas([[0, [0-19, 1-19, 2-10],[0,1,2,3,4,5,6]], [1, [4-25],[0,1,2]]], [0-15, 1-10, 2-3, 3-10,4-1], [0-1,1-1,2-2,3-3,4-4], Lucro, Solucao).
-escola_de_linguas([[0, [0-10],[0,1]]], [0-10], [0-20], Lucro, Solucao).
+%escola_de_linguas([[0, [0-19, 1-19, 2-10,3-4,4-5],[0,1,2,3,4,5,6]], [1, [4-25],[0,1,2]]], [0-15, 1-10, 2-3, 3-10,4-1], [0-1,1-1,2-2,3-3,4-4], Lucro, Solucao).
+escola_de_linguas([[0, [0-10,1-10],[0,1]], [1, [0-5],[0,1]],[2, [0-5],[0,1]]], [0-10,1-10], [0-20,1-40], Lucro, Solucao).
 
 
 falha(N):-N in 0..10.

### 2015-12-09 proj2/algumasconsideracoessobreoproblemaemestudo.pl
mais restriçoes

@@ -16,7 +16,7 @@
 	% 2 - minimo 2 horas por semana/linguas
 	% 3 - maximo 4 horas/semana/lingua
 	% 4 - 15 alunos/lingua
-	% 5 - em cada dia nao pode haver mais do que x aulas em cada intervalo inicio-fim, para todos os intervalos
+	% 5 - em cada dia nao pode haver mais do que x aulas
 
 clearScreen:-write('\e[2J').
 
@@ -79,8 +79,8 @@ labelAll([],_).
 labelAll(Sol, Cost):-
 	labelAll(Sol, Cost, []).
 
-labelAll([], Cost, Flat):-write('gona label'),nl, maximize(labeling([], Flat), Cost).
-labelAll([Indice-[Dia1,Dia2,Dia3,Dia4,Dia5]|Tail], Cost, Flat):-
+labelAll([], Cost, Flat):-write('gona label'),nl, labeling([maximize(Cost)], Flat).
+labelAll([_-[Dia1,Dia2,Dia3,Dia4,Dia5]|Tail], Cost, Flat):-
 	write('labeling'),nl,
 	append(Flat, Dia1, Flat1),
 	append(Flat1, Dia2, Flat2),
@@ -96,7 +96,7 @@ restrict_teacher_hours(Solucao, [Professor|T]):-
 	restrict_teacher_hours(Solucao, T).
 
 %
-restrict_teacher([], _, Acum):- Acum #< 15, write(Acum).
+restrict_teacher([], _, Acum):- Acum #< 15.
 restrict_teacher([H|T], Indice, Acum):-
 	H = _-Dias,
 	teacher_times(Dias, Indice, Time),
@@ -111,55 +111,69 @@ teacher_times([[_, Horas, ProfessorResponsavel]|Dias], Indice, Time):-
 	teacher_times(Dias, Indice, TempoRestante),
 	Time #= TempoQueOProfDaEstaAula+TempoRestante.
 
+restrict_course_times([]).
+restrict_course_times([_-[[_,H1,_],[_,H2,_],[_,H3,_],[_,H4,_],[_,H5,_]]|T]):-
+	sum([H1,H2,H3,H4,H5],#=<,4),
+	sum([H1,H2,H3,H4,H5],#>=,2),
+	restrict_course_times(T).
+	
 
+	
 
 
 escola_de_linguas(Professores, Candidaturas, Linguas, Lucro, Solucao):-
 	length(Professores, NProfessores),
[…]

### 2015-12-10 proj2/algumasconsideracoessobreoproblemaemestudo.pl
tabela

@@ -1,5 +1,6 @@
 :-use_module(library(clpfd)).
 :-use_module(library(aggregate)).
+:-use_module(library(lists)).
 
 
 
@@ -91,48 +92,43 @@ labelAll([_-[Dia1,Dia2,Dia3,Dia4,Dia5]|Tail], Cost, Flat):-
 
 restrict_teacher_hours(_, []).
 restrict_teacher_hours(Solucao, [Professor|T]):-
-	Professor=[Indice,_,_],
-	restrict_teacher(Solucao, Indice, 0),
-	restrict_teacher_hours(Solucao, T).
-
-%
-restrict_teacher([], _, Acum):- Acum #< 15.
-restrict_teacher([H|T], Indice, Acum):-
-	H = _-Dias,
-	teacher_times(Dias, Indice, Time),
-	Acum1 #= Time+Acum,
-	restrict_teacher(T, Indice, Acum1).
-
-
-teacher_times([], _, 0).
-teacher_times([[_, Horas, ProfessorResponsavel]|Dias], Indice, Time):-
-	ProfessorResponsavel #= Indice #<=> B,
-	TempoQueOProfDaEstaAula #= B*Horas,
-	teacher_times(Dias, Indice, TempoRestante),
-	Time #= TempoQueOProfDaEstaAula+TempoRestante.
-
-restrict_course_times([]).
-restrict_course_times([_-[[_,H1,_],[_,H2,_],[_,H3,_],[_,H4,_],[_,H5,_]]|T]):-
-	sum([H1,H2,H3,H4,H5],#=<,4),
-	sum([H1,H2,H3,H4,H5],#>=,2),
-	restrict_course_times(T).
-	
 
 	
-
-
+generate_domain_day_list([],[],[]).
+generate_domain_day_list([[_,Hora,Prof]|Dias], H, P):-
+	generate_domain_day_list(Dias, HRest, PRest),
+	append([Hora], HRest, H),
+	append([Prof], PRest, P).
+	
+generate_variable_domain_list([],[],[]).
+generate_variable_domain_list([_-Dias|Solucoes], H, P):-
+	generate_domain_day_list(Dias, HThis, PThis),
+	generate_variable_domain_list(Solucoes, HRest, PRest),
+	append(HThis, HRest, H),
+	append(PThis, PRest, P).	
+	
 escola_de_linguas(Professores, Candidaturas, Linguas, Lucro, Solucao):-
 	length(Professores, NProfessores),
 	length(Linguas, NLinguas),
[…]

### 2015-12-11 proj2/algumasconsideracoessobreoproblemaemestudo.pl
ITS A MIRACLE

@@ -35,28 +35,50 @@ professoresQueDaoLingua(Lingua,  [ [_, _,_] | OutrosProfs], Lista):-
 testeProfs(Sol):-
 	professoresQueDaoLingua(5, [[0, [0-1,1-2],_], [1, [0-2],_], [2, [3-2],_]], Sol).
 
+generate_professor_availability([], []).
+generate_professor_availability([[Indice,_,Dias]|Professors], Table):-
+	generate_professor_availability(Dias,Indice,TableThis),
+	generate_professor_availability(Professors, TableRest),
+	append(TableThis, TableRest, TableN),
+	append([[-1,0],[-1,1],[-1,2],[-1,3],[-1,4],[-1,5]], TableN, Table).
+
+generate_professor_availability([], _, []).
+generate_professor_availability([Dia|Dias],Indice, Table):-
+	generate_professor_availability(Dias, Indice, TableRest),
+	append([[Indice,Dia]], TableRest, Table).
 
 
 
-initialize_days_list([],_,_,_).
-initialize_days_list([H|T], N, NProfessores, FdSetDeProfs):-
+
+
+initialize_days_list([],_,_,_,_).
+initialize_days_list([H|T], N, NProfessores, FdSetDeProfs, ProfAvailable):-
 	H = [N,Hora,Prof],
-
-	Prof in_set FdSetDeProfs,
+	Prof #= -1 #=> Hora #= 0,
+	table([[Prof,N]],ProfAvailable),
+	Prof in_set FdSetDeProfs ,
 	domain([Hora], 0, 4),
 	N1 is N+1,
-	initialize_days_list(T, N1, NProfessores, FdSetDeProfs).
+	initialize_days_list(T, N1, NProfessores, FdSetDeProfs, ProfAvailable).
 
 
-initialize_solution([], _,_,_).
-initialize_solution([H|T], Lingua, NProfessores, Professores):-
+
+
+initialize_solution([], _,_,_,_).
+initialize_solution([H|T], Lingua, NProfessores, Professores, ProfAvailable):-
 	H = Lingua-Lista,
 	length(Lista, 5),
 	professoresQueDaoLingua(Lingua, Professores, ListaDeProfs),
 	list_to_fdset([-1|ListaDeProfs], FdSetDeProfs),
[…]

### 2015-12-12 proj2/algumasconsideracoessobreoproblemaemestudo.pl
restriçao das salas

@@ -1,9 +1,6 @@
 :-use_module(library(clpfd)).
 :-use_module(library(aggregate)).
 :-use_module(library(lists)).
-
-
-
 %entradas
 	% professor - (indiceP-[Lista de linguas-rendimentoporhora|_]-[Lista de dias|_])
 	% candidatura - (Lingua-Numero)
@@ -19,42 +16,12 @@
 	% 4 - 15 alunos/lingua
 	% 5 - em cada dia nao pode haver mais do que x aulas
 
-clearScreen:-write('\e[2J').
-
-
-professoresQueDaoLingua(_, [], []).
-professoresQueDaoLingua(Lingua,  [ [Prof, Linguas,_] | OutrosProfs], Lista):-
-	member(Lingua-_, Linguas),
-	professoresQueDaoLingua(Lingua, OutrosProfs, Lista2),
-	append([Prof], Lista2, Lista).
-
-professoresQueDaoLingua(Lingua,  [ [_, _,_] | OutrosProfs], Lista):-
-	professoresQueDaoLingua(Lingua, OutrosProfs, Lista).
-
-
-testeProfs(Sol):-
-	professoresQueDaoLingua(5, [[0, [0-1,1-2],_], [1, [0-2],_], [2, [3-2],_]], Sol).
-
-generate_professor_availability([], []).
-generate_professor_availability([[Indice,_,Dias]|Professors], Table):-
-	generate_professor_availability(Dias,Indice,TableThis),
-	generate_professor_availability(Professors, TableRest),
-	append(TableThis, TableRest, TableN),
-	append([[-1,0],[-1,1],[-1,2],[-1,3],[-1,4],[-1,5]], TableN, Table).
-
-generate_professor_availability([], _, []).
-generate_professor_availability([Dia|Dias],Indice, Table):-
-	generate_professor_availability(Dias, Indice, TableRest),
-	append([[Indice,Dia]], TableRest, Table).
-
-
-
-
+%%%%% INICIALIZAÇÂO %%%%%
 
 initialize_days_list([],_,_,_,_).
 initialize_days_list([H|T], N, NProfessores, FdSetDeProfs, ProfAvailable):-
 	H = [N,Hora,Prof],
-	Prof #= -1 #=> Hora #= 0,
[…]

### 2015-12-12 proj2/algumasconsideracoessobreoproblemaemestudo.pl
fixed maximization issue

@@ -1,6 +1,9 @@
 :-use_module(library(clpfd)).
 :-use_module(library(aggregate)).
 :-use_module(library(lists)).
+
+
+
 %entradas
 	% professor - (indiceP-[Lista de linguas-rendimentoporhora|_]-[Lista de dias|_])
 	% candidatura - (Lingua-Numero)
@@ -16,12 +19,42 @@
 	% 4 - 15 alunos/lingua
 	% 5 - em cada dia nao pode haver mais do que x aulas
 
-%%%%% INICIALIZAÇÂO %%%%%
+clearScreen:-write('\e[2J').
+
+
+professoresQueDaoLingua(_, [], []).
+professoresQueDaoLingua(Lingua,  [ [Prof, Linguas,_] | OutrosProfs], Lista):-
+	member(Lingua-_, Linguas),
+	professoresQueDaoLingua(Lingua, OutrosProfs, Lista2),
+	append([Prof], Lista2, Lista).
+
+professoresQueDaoLingua(Lingua,  [ [_, _,_] | OutrosProfs], Lista):-
+	professoresQueDaoLingua(Lingua, OutrosProfs, Lista).
+
+
+testeProfs(Sol):-
+	professoresQueDaoLingua(5, [[0, [0-1,1-2],_], [1, [0-2],_], [2, [3-2],_]], Sol).
+
+generate_professor_availability([], []).
+generate_professor_availability([[Indice,_,Dias]|Professors], Table):-
+	generate_professor_availability(Dias,Indice,TableThis),
+	generate_professor_availability(Professors, TableRest),
+	append(TableThis, TableRest, TableN),
+	append([[-1,0],[-1,1],[-1,2],[-1,3],[-1,4],[-1,5]], TableN, Table).
+
+generate_professor_availability([], _, []).
+generate_professor_availability([Dia|Dias],Indice, Table):-
+	generate_professor_availability(Dias, Indice, TableRest),
+	append([[Indice,Dia]], TableRest, Table).
+
+
+
+
 
 initialize_days_list([],_,_,_,_).
 initialize_days_list([H|T], N, NProfessores, FdSetDeProfs, ProfAvailable):-
 	H = [N,Hora,Prof],
-	Prof #= -1 #<=> Hora #= 0,
[…]

### 2015-12-13 proj2/algumasconsideracoessobreoproblemaemestudo.pl
fix problem with teacher hours restriction

@@ -114,13 +114,13 @@ labelAll([_-[Dia1,Dia2,Dia3,Dia4,Dia5]|Tail], Cost, Flat):-
 restrict_teacher_hours(_,_,[], _).
 restrict_teacher_hours(Horas, ProfessoresDominio, [[Indice,_,_]|T], NProfs):-
 	make_prof_table(Indice, Table, NProfs),!,
-	restrict_teacher(Table, ProfessoresDominio, Rest),
+	append(Table, [[-1, 0]], NTable),
+	restrict_teacher(NTable, ProfessoresDominio, Rest),
 	sum_product(Horas, Rest, Sum),
 	Sum #=< 15,
 	restrict_teacher_hours(Horas, ProfessoresDominio, T, NProfs).
 
 restrict_teacher(_, [], []).
-restrict_teacher(Table, [-1|Ps],[0|T]):- restrict_teacher(Table, Ps, T).
 restrict_teacher(Table, [P|Ps], [Rest|T]):-
 	table([[P,Rest]], Table),
 	restrict_teacher(Table, Ps, T).
@@ -131,6 +131,14 @@ sum_product([X|Xs],[Y|Ys], Sum):-
 	sum_product(Xs,Ys,SumRest),
 	Sum #= X*Y + SumRest.
 
+make_prof_table(Indice, Lista, NProfs):-
+	Length is NProfs+1,
+	length(Lista, Length),
+	iterate_prof_table(Lista, NProfs, Indice).
+
+iterate_prof_table(_,-1,_).
+iterate_prof_table(Lista, NProfs, NProfs):- !, nth0(NProfs, Lista, [NProfs, 1]), Next is NProfs-1, iterate_prof_table(Lista, Next, NProfs).
+iterate_prof_table(Lista, NProfs, Indice):- nth0(NProfs, Lista, [NProfs,0]), Next is NProfs-1, iterate_prof_table(Lista, Next, Indice).
 
 
 
@@ -180,19 +188,12 @@ escola_de_linguas(Professores, Candidaturas, Linguas, Rooms, Lucro, Solucao):-
 	fd_statistics.
 
 
-make_prof_table(Indice, Lista, NProfs):-
-	Length is NProfs+1,
-	length(Lista, Length),
-	iterate_prof_table(Lista, NProfs, Indice).
 
[…]

### 2015-12-13 proj2/algumasconsideracoessobreoproblemaemestudo.pl
Milagre de natal

@@ -54,10 +54,12 @@ generate_professor_availability([Dia|Dias],Indice, Table):-
 initialize_days_list([],_,_,_,_).
 initialize_days_list([H|T], N, NProfessores, FdSetDeProfs, ProfAvailable):-
 	H = [N,Hora,Prof],
-	Prof #= -1 #=> Hora #= 0,
-	table([[Prof,N]],ProfAvailable),
-	Prof in_set FdSetDeProfs ,
-	domain([Hora], 0, 4),
+	Prof #= -1 #<=> Hora #= 0,
+	findall(ProfA, member( [ProfA, N], ProfAvailable), ListaDeProfsA),
+	list_to_fdset(ListaDeProfsA, FdSetDisponivel),
+	fdset_intersection(FdSetDisponivel, FdSetDeProfs, FdSetFinal),
+	Prof in_set FdSetFinal,
+	domain([Hora], 0, 2),
 	N1 is N+1,
 	initialize_days_list(T, N1, NProfessores, FdSetDeProfs, ProfAvailable).
 
@@ -77,39 +79,15 @@ initialize_solution([H|T], Lingua, NProfessores, Professores, ProfAvailable):-
 	restrict_course_times([]).
 	restrict_course_times([_-[[_,H1,_],[_,H2,_],[_,H3,_],[_,H4,_],[_,H5,_]]|T]):-
 		sum([H1,H2,H3,H4,H5],#=<,4),
-		sum([H1,H2,H3,H4,H5],#>=,2),
+		sum([H1,H2,H3,H4,H5],#\=,1),
+		sum([H1,H2,H3,H4,H5],#>=,0),
 		restrict_course_times(T).
 
-forceEmpty([]).
-forceEmpty([_-Horas-Prof|T]):-
-	Horas in {0},
-	Prof in {-1},
-	forceEmpty(T).
 
-excludeEmpty([],_).
-excludeEmpty([Lingua-Horarios|Solucoes], Candidaturas):-
-	member(Lingua-0, Candidaturas),
-	forceEmpty(Horarios),
-	excludeEmpty(Solucoes, Candidaturas).
+% [ 0-[ [0, 2, 4], [1, 3, 5] ], 1-[[0,4,5], [1,3,5],[2,6,6] ]  ] labeling([ffc,bisect,up,maximize(Cost)], Flat).
 
-	excludeEmpty([_-_|Solucoes], Candidaturas):-
[…]

### 2015-12-13 proj2/doc/article.tex
report under way

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage[utf8]{inputenc}
%
\begin{document}

\tableofcontents
%
\mainmatter              % start of the contributions
%
\title{Solving an optimization problem employing restraint programming}
%
\titlerunning{Restraint Programming}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{João Cabral\inst{1} \and João Mota\inst{2}}
%
\authorrunning{João Cabral \and João Mota} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{João Cabral, João Mota}
%
\institute{Faculdade de Engenharia da Universidade do Porto, Porto, Rua Roberto Frias s/n 4200-465, Portugal,\\
\email{(up201303462, up201304395)@fe.up.pt}}

\maketitle              % typeset the title of the contribution

\begin{abstract}
This paper goes over the theoretical and practical aspects of the resolution of an optimization problem,
namely the synthesis of a schedule for a language school, employing restraint programming, more precisely, 
SICSTUS Prolog's restraint programming capabilities.
The paper contains a description of the problem, a detailed analysis of the aproach used to solve it, 
an explanation of how the solution should be interpreted, a description of obtained results and, finally, conclusions.
%n ta bem isto...

\keywords{optimization, restraint programming, PROLOG}
\end{abstract}
%
\section{Introduction}
Within the context of the Logic Programming course of MIEIC, the authors of this paper were asked to solve an optimization problem in the most efficient way possible using the restraint programming capabilities of SICStus PROLOG. The optimization problem under study is that of producing a working schedule for a language school that maximizes it's (monetary) gains,  given a number of inputs of variable length, aswell as a number of restrictions of various types. The rest of the paper is structured as follows:

\begin{itemize}
	\item Problem Description - A detailed description of the problem that's under study in this paper.
	\item Approach - The authors' approach to the modelization of the problem as a Constraint Satisfaction Problem (henceforth called CSP)
	\item Solution Presentation - Explanation of the predicades that allow a user to interpret the output of the implemented solution.
	\item Results - Examples of different instances of the problem with differing complexities, aswell as a discussion of the temporal and spatial efficiency of the implemented solution.
	\item Conclusions and Future Work - Discussion of the conclusions to withdraw from the obtained results.
	\item References - Any and all work the authors' used to further the completion of the proposed task.
	\item Annex - Source code, Results, Data files regarding the program's efficiency
\end{itemize}

\section{Problem Description}
As was mentioned in the introduction, the optimization problem under study in this paper is that of producing a working schedule for a language school that maximizes it's gains, given a number of inputs of varialbe length, coupled with a number of restricitons.  The inputs consist of a list of teachers, a list of registrations for the language courses, a list of language courses provided and the number of rooms available at the hypothetical school. As for restrictions, they are as follows:

\begin{itemize}
	\item A student will pay an hourly amount for the attendance of a course, which will differ based on the course.
	\item A teacher can work at most 15 hours per week, and will only work on specific days
	\item A teacher will earn a certain amount per hour, based on the language taught. If the teacher is working more than 10 hours per week, the earnings for the excess will be decreased by 20\%.
	\item Each language course, if taught, will have no less than 2 weekly hours, and no more than 4.
	\item Each course can be frequented by at most 15 students
	\item The school has a finite number of classrooms
\end{itemize}

There are, of course, other, implicit, restrictions, such as the fact that a teacher cannot teach a language course for a language that he's not qualified to teach, or the fact that a course cannot be taught if there are no students registered for it, or if there's no qualified teachers to teach it.

\section{Approach}
In this section, the more technical aspects of solving the problem are discussed, such as the modelization of the problem as a CSP and the implementation of its restrictions. It also goes over the search strategies employed in the program's labeling phase.

\subsection{Problem Modelization}
In order to function, the program implemented by the paper's authors to solve the problem unser study needs to be fed a number of inputs. As was mentioned before, these consist of a list of teachers, a list of course registrations, the list of available courses and the number of available rooms. The list of teachers is implemented as follows:
\begin{verbatim}
	[[TeacherIndex, [LanguageIndex-PayPerHour|OtherLanguages], [DayAvailable|OtherDays] ]
	| OtherTeachers]
\end{verbatim}

Where the first list within a teacher represents the languages that teacher's qualified to teach (and it's hourly pay) and the second represents the indexes of the days a teacher's available to teach.

The course registrations are implemented as follows:
\begin{verbatim}
	[LanguageIndex-NumberOfRegistrations|OtherLanguages]
\end{verbatim}

The available courses list is implemented as follows: 
\begin{verbatim}
	[LanguageIndex-HourlyCost|OtherLanguages]
\end{verbatim}

The number of rooms in this problem is interpreted by the paper's authors as the number of classes that can be taught simultaneously each day. This is done for the sake of simplifying the problem, otherwise the solution would have to take into account the concept of hours and different classes for the same course.

Given all of these inputs, the program will produce a solution that's formatted as follows:
\begin{verbatim}
	[LanguageIndex-[[DayIndex, HoursTaught, TeacherIndex|OtherDays]]|OtherLanguages]
\end{verbatim}

Where the list attached to every LanguageIndex consists of a listing of the hours of class taught each day of the week, and the teacher that's assigned to teach them.

\subsection{Decision Variables}
As was mentioned in the section above, the solution is formatted as follows:
\begin{verbatim}
	[LanguageIndex-[[DayIndex, HoursTaught, TeacherIndex|OtherDays]]|OtherLanguages]
\end{verbatim}

In the problem under study, the decision variables are HoursTaught and TeacherIndex, LanguageIndex and DayIndex being fixed in order to be able to present the solution in a coherent, comprehensive way.

The domain for HoursTaught is 0-4, seeing as the most hours of class a course can have per week is 4. The domain for TeacherIndex, however, is rather complex. It'll be the list of  TeacherIndexes with hours to spare (from the weekly 15 they are alloted), who are qualified to teach the language in question and are not teaching any other language that day (this is due to the way the concept of classrooms is interpreted in the context of this solution, which is explained in the section above this one). In the particular case of courses that didn't open for lack of students or teachers, the TeacherIndex variable will assume the value -1.

\subsection{Restraints}
% FALTA BOTAR AQUI CENAS
The restraints for this particular problem are listed in section 2. As such this section will focus mainly on restrictions' implementation aspects. All restricitions concerning monetary values are approached during the profit calculating phase of the solution, which is further explained in the next section of this paper. For the restriction concerning the number of hours of class a certain course can have per week, the program will gather the values for the hours taught each day and say that their sum must be grater than or equal to 2 and lesser than or equal to 4.

\begin{verbatim}
restrict_course_times([]).
	restrict_course_times([_-[[_,H1,_],[_,H2,_],[_,H3,_],[_,H4,_],[_,H5,_]]|T]):-
		sum([H1,H2,H3,H4,H5],#=<,4),
		sum([H1,H2,H3,H4,H5],#>=,2),
		restrict_course_times(T).
\end{verbatim}

There are, however, exceptions to this restriction in the authors' implementation of a solution, where a course can have 0 weekly hours taught when it doesn't have anyone to attend or teach it (effectively signifying that this course didn't open).

Concerning the restriction that a teacher can teach at most 15 hours per week,

\subsection{Evaluation Function}
Since the program has the goal of maximizing the monetary gains for the school, the evaluation function for the problem at hand is the total weekly gains of the school. The value for the total weekly gains of the school is calculated by summing the daily gains for each language, which can be calculated using the following formula:

\begin{equation}
	profit = hoursTaught*((students*pricePerHour)+(-teacherSalary))
\end{equation}

The predicates below are used to calculate the profit:
%ATUALIZAR AQUI SE ISTO FOR MUDADO
\begin{verbatim}
make_profit([], _, _, _,_, 0).
make_profit([Solucao|Solucoes], Linguas, Professores,TabelaCustos, Candidaturas, Lucro):-
	make_profit_language(Solucao, Linguas, Professores, TabelaCustos, Candidaturas, LucroLingua),
	make_profit(Solucoes, Linguas, Professores, TabelaCustos, Candidaturas, LucroResto),
	Lucro #= LucroResto+LucroLingua.

make_profit_language(Lingua, Linguas, Professores, TabelaCustos, 
Candidaturas, LucroLingua):-
	Lingua = IndiceL-Dias,
	make_profit_day(Dias, Linguas, Professores, TabelaCustos, 
	Candidaturas, IndiceL, LucroLingua).
make_profit_day([], _, _,_, _, _, 0).
make_profit_day([Dia|OutrosDias], Linguas, Professores, TabelaCustos,
 Candidaturas, IndiceL, LucroDia):-
	Dia = [_, Horas, Professor],
	table([[Professor, IndiceL, Remuneracao]],TabelaCustos),
	Prejuizo #= Remuneracao * Horas,
	member(IndiceL-Preco, Linguas),
	member(IndiceL-NumeroCandidatos, Candidaturas),

	Lucro #= Preco*NumeroCandidatos*Horas,
	LucroDiaAtual #= (Lucro-Prejuizo),
	make_profit_day(OutrosDias, Linguas, Professores,TabelaCustos, 
	Candidaturas, IndiceL, LucroRest),
	LucroDia #= LucroRest + LucroDiaAtual.
\end{verbatim}

Then, in the labeling phase, the program is told to maximize Profit, as seen here:

\begin{verbatim}
labeling([ff,enum,down,maximize(Profit)], Flat).
\end{verbatim}

\subsection{Search Strategy}

%
% ---- Bibliography ----
%
\begin{thebibliography}{5}
%
\bibitem {clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and
boundary-value problems for Hamiltonian systems.
Arch. Rat. Mech. Anal. 78, 315--333 (1982)

\bibitem {clar:eke:2}
Clarke, F., Ekeland, I.:
Solutions p\'{e}riodiques, du
p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
Note CRAS Paris 287, 1013--1015 (1978)

\bibitem {mich:tar}
Michalek, R., Tarantello, G.:
Subharmonic solutions with prescribed minimal
period for nonautonomous Hamiltonian systems.
J. Diff. Eq. 72, 28--55 (1988)

\bibitem {tar}
Tarantello, G.:
Subharmonic solutions for Hamiltonian
systems via a $\bbbz_{p}$ pseudoindex theory.
Annali di Matematica Pura (to appear)

\bibitem {rab}
Rabinowitz, P.:
On subharmonic solutions of a Hamiltonian system.
Comm. Pure Appl. Math. 33, 609--633 (1980)

\end{thebibliography}

\end{document}

### 2015-12-13 proj2/doc/article.tex
added a lil bit moar

@@
 
-Concerning the restriction that a teacher can teach at most 15 hours per week,
+Concerning the restriction that a teacher can teach at most 15 hours per week, a number of predicates are in place that process and convert the input into formats that make it easy to gather the variables that concern the number of hours a teacher teaches per week.
+It is then said that the sum of these variables cannot exceed 15 (the code concerning this can be viewed in the annex, and will not be replicated here for reasons of spatial economy).
 

### 2015-12-13 proj2/algumasconsideracoessobreoproblemaemestudo.pl
Print formatting and some other stuff

@@ -59,7 +59,7 @@ initialize_days_list([H|T], N, NProfessores, FdSetDeProfs, ProfAvailable):-
 	list_to_fdset(ListaDeProfsA, FdSetDisponivel),
 	fdset_intersection(FdSetDisponivel, FdSetDeProfs, FdSetFinal),
 	Prof in_set FdSetFinal,
-	domain([Hora], 0, 2),
+	Hora in {0} \/{2},
 	N1 is N+1,
 	initialize_days_list(T, N1, NProfessores, FdSetDeProfs, ProfAvailable).
 
@@ -84,9 +84,8 @@ initialize_solution([H|T], Lingua, NProfessores, Professores, ProfAvailable):-
 		restrict_course_times(T).
 
 
-% [ 0-[ [0, 2, 4], [1, 3, 5] ], 1-[[0,4,5], [1,3,5],[2,6,6] ]  ] labeling([ffc,bisect,up,maximize(Cost)], Flat).
 
-labelAll(H,P,Cost):- append(H, P, List), labeling([all,down,maximize(Cost)], List).
+labelAll(H,P,Cost):- append(H, P, List), labeling([down,leftmost,time_out(30000,_),maximize(Cost)], List).
 
 
 restrict_teacher_hours(_,_,[], _).
@@ -134,7 +133,6 @@ generate_variable_domain_list([_-Dias|Solucoes], H, P):-
 custo_nulo([], 0).
 custo_nulo(TabelaCustos, NLinguas):-
 	NLinguas \= 0,
-	write(NLinguas),
 	N is NLinguas-1,
 	custo_nulo(TabelaCustosR, N),
 	append([[-1,N,0]], TabelaCustosR, TabelaCustos).
@@ -145,9 +143,8 @@ escola_de_linguas(Professores, Candidaturas, Linguas, Rooms, Lucro, Solucao):-
 	length(Solucao, NLinguas),
 	generate_professor_availability(Professores, ProfAvailable),
 	cria_tabela_custos(Professores, TabelaCustosReais, []),
-	write(NLinguas), nl, nl,
 	custo_nulo(CustoNulo, NLinguas),
-	append(CustoNulo, TabelaCustosReais, TabelaCustos),write(TabelaCustos),
[…]

### 2015-12-13 proj2/doc/article.tex
temp

@@
 \subsection{Search Strategy}
+ NUM SEI OQ HEI-DE POR AQUI
+
+ \section{Solution Presentation}
+ After the labeling phase, the solution is presented in a way that isn't easily legible by users who have no knowledge of the program's implementation.
+To aid in the observation and interpretation of the solution, the authors created a predicate which converts it to a more understandable format.
 

### 2015-12-13 proj2/algumasconsideracoessobreoproblemaemestudo.pl

@@ -205,7 +205,8 @@ make_profit_day([Dia|OutrosDias], Linguas, Professores, TabelaCustos, Candidatur
 	member(IndiceL-Preco, Linguas),
 	member(IndiceL-NumeroCandidatos, Candidaturas),
 
-	Lucro #= Preco*NumeroCandidatos*Horas,
+	(NumeroCandidatos < 15, Lucro #= Preco*NumeroCandidatos*Horas);
+	(NumeroCandidatos >= 15, Lucro #= Preco*15*Horas),
 	LucroDiaAtual #= (Lucro-Prejuizo),
 	make_profit_day(OutrosDias, Linguas, Professores,TabelaCustos, Candidaturas, IndiceL, LucroRest),
 	LucroDia #= LucroRest + LucroDiaAtual.

### 2015-12-13 proj2/doc/article.tex
report almost done

@@
 \begin{document}
-
 \tableofcontents
 %
-\mainmatter              % start of the contributions
+%\mainmatter              % start of the contributions
 %
@@
 %
+ISTO TA A FICAR APERTADO D PAGS, SE EXCEDER TIRA-SE OS VERBATIMS TODOS E OS PROFS PODEM IR LER AOS ANEXOS Q SE FODEM
 \section{Introduction}
@@
 To aid in the observation and interpretation of the solution, the authors created a predicate which converts it to a more understandable format.
+Here's an example of a possible solution:
+
+ \begin{verbatim}
+ Language 0 Time Table
+ _______________
+Language 1 Time Table
+_______________
+Language 2 Time Table
+_______________
+Language 3 Time Table
+_______________
+Monday - 2 hours with professor 0
+Tuesday - 2 hours with professor 0
+
+Language 4 Time Table
+_______________
+Wednesday - 2 hours with professor 4
+Friday - 2 hours with professor 0
+
+Language 5 Time Table
+_______________
+Thursday - 2 hours with professor 3
+
+O lucro total obtido e de 994
+Resumptions: 967755
+Entailments: 124262
+Prunings: 348556
+Backtracks: 1566
+Constraints created: 749
+O resultado foi obtido em 0.140 segundos.
+ \end{verbatim}
+
+ As can be seen above, a time table is created for each language in which only the days in which the class is taught are displayed, along with the number of hours taught that day and the teacher who teaches them.
+This is achieved via the creation of 5 predictes that make the connection between the index of days and the actual day, coupled with predicates that format and write the information on the screen.
+(This code that does this can be viewed in the annex, namely in the predicate printSolution).
+
+ \section{Results}
+ %BOTA GRAFICOS E TAL
+
+ \section{Conclusions and Future Work}
+ Having implemented a solution to the problem under study, the paper's authors are able to appreciate the advantages of employing restraint programming in the solving of problems that are similar to the one studied here.
+Indeed, the implemented solution shows that the use of this technology allows for a relatively simple implementation of a solution, when compared to possible implementations using other technologies.
+%aspetos a melhorar???
 

### 2015-12-13 proj2/algumasconsideracoessobreoproblemaemestudo.pl

@@ -204,7 +204,7 @@ make_profit_day([Dia|OutrosDias], Linguas, Professores, TabelaCustos, Candidatur
 
 	Inscritos is min(NumeroCandidatos, 15),
 	Lucro #= Preco*Inscritos*Horas,
-s
+
 	LucroDiaAtual #= (Lucro-Prejuizo),
 	make_profit_day(OutrosDias, Linguas, Professores,TabelaCustos, Candidaturas, IndiceL, LucroRest),
 	LucroDia #= LucroRest + LucroDiaAtual.

### 2015-12-13 proj2/doc/article.tex
added bibliography

@@
 %
-\begin{thebibliography}{5}
+\begin{thebibliography}{2}
 %
-\bibitem {clar:eke}
-Clarke, F., Ekeland, I.:
-Nonlinear oscillations and
-boundary-value problems for Hamiltonian systems.
-Arch. Rat. Mech. Anal. 78, 315--333 (1982)
 
 \bibitem {clar:eke:2}
-Clarke, F., Ekeland, I.:
-Solutions p\'{e}riodiques, du
-p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
-Note CRAS Paris 287, 1013--1015 (1978)
+SICStus Prolog documentation:
+https://sicstus.sics.se/documentation.html
 
-\bibitem {mich:tar}
-Michalek, R., Tarantello, G.:
-Subharmonic solutions with prescribed minimal
-period for nonautonomous Hamiltonian systems.
-J. Diff. Eq. 72, 28--55 (1988)
-
-\bibitem {tar}
-Tarantello, G.:
-Subharmonic solutions for Hamiltonian
-systems via a $\bbbz_{p}$ pseudoindex theory.
-Annali di Matematica Pura (to appear)
-
-\bibitem {rab}
-Rabinowitz, P.:
-On subharmonic solutions of a Hamiltonian system.
-Comm. Pure Appl. Math. 33, 609--633 (1980)
+\bibitem{ster:ling}
+Leon, S., Sterling:
+The Art of Prolog
 

### 2015-12-13 proj2/algumasconsideracoessobreoproblemaemestudo.pl

@@ -85,7 +85,7 @@ initialize_solution([H|T], Lingua, NProfessores, Professores, ProfAvailable):-
 
 
 
-labelAll(H,P,Cost):- append(H, P, List), labeling([down,leftmost,time_out(30000,_),maximize(Cost)], List).
+labelAll(H,P,Cost):- append(H, P, List), labeling([up,bisect,ff,timeout(60000,),maximize(Cost)], List).
 
 
 restrict_teacher_hours(_,_,[], _).
@@ -164,7 +164,9 @@ escola_de_linguas(Professores, Candidaturas, Linguas, Rooms, Lucro, Solucao):-
 teste:-
 	statistics(runtime, [T0| _]),
 	%escola_de_linguas([[0, [0-0,1-0,3-0,4-0,5-0,6-0],[0,1,2,3,4]], [1, [0-0,2-0],[0,1,2,3,4]],[2, [0-0,3-0,4-0,5-0,6-0],[0,1,2,3,4]],[3, [0-0,1-0,3-0,4-0,5-0,6-0],[0,1,2,3,4]]], [0-10,1-10,2-10,3-10,4-10,5-10,6-10], [0-1,1-1,2-1,3-1,4-1,5-1,6-1], 40,Lucro, Solucao),
-	escola_de_linguas([[0, [0-1, 1-1,2-1,3-0,4-0],[0,1,4,5,6]], [1, [0-1, 1-1,2-1],[0,4]], [2, [0-1, 1-1,3-1],[0,3]],[3, [0-1, 1-1,4-1,5-2,6-1],[0,2,3,4,5,6]],[4, [0-0	, 1-1,4-1],[0,1,2,3,4,5,6]]], [0-15, 1-10,2-10,3-10,4-10,5-10], [0-1,1-1,2-1,3-10,4-10,5-10], 1,_, _),
+	escola_de_linguas([[0, [0-1, 1-1,2-1,3-0,4-0],[0,1,4,5,6]], [1, [0-1, 1-1,2-1],[0,4]], [2, [0-1, 1-1,3-1],[0,3]] ],
+	[0-15, 1-10,2-10,3-10,4-10,5-10],
+	[0-1,1-1,2-1,3-10,4-10,5-10], 4,_, _),
 	statistics(runtime, [T1|_]),
 	T is T1 - T0,
 	format('O resultado foi obtido em ~3d segundos.~n', [T]).

### 2015-12-13 proj2/doc/article.tex
source code in the annex

@@
 
+\section{Annex}
+
+\subsection{Source code}
+
+\begin{verbatim}
+:-use_module(library(clpfd)).
+:-use_module(library(aggregate)).
+:-use_module(library(lists)).
+
+clearScreen:-write('\e[2J').
+
+professoresQueDaoLingua(_, [], []).
+professoresQueDaoLingua(Lingua,  [ [Prof, Linguas,_] | OutrosProfs], Lista):-
+	member(Lingua-_, Linguas),
+	professoresQueDaoLingua(Lingua, OutrosProfs, Lista2),
+	append([Prof], Lista2, Lista).
+
+professoresQueDaoLingua(Lingua,  [ [_, _,_] | OutrosProfs], Lista):-
+	professoresQueDaoLingua(Lingua, OutrosProfs, Lista).
+
+generate_professor_availability([], []).
+generate_professor_availability([[Indice,_,Dias]|Professors], Table):-
+	generate_professor_availability(Dias,Indice,TableThis),
+	generate_professor_availability(Professors, TableRest),
+	append(TableThis, TableRest, TableN),
+	append([[-1,0],[-1,1],[-1,2],[-1,3],[-1,4],[-1,5]], TableN, Table).
+
+generate_professor_availability([], _, []).
+generate_professor_availability([Dia|Dias],Indice, Table):-
+	generate_professor_availability(Dias, Indice, TableRest),
+	append([[Indice,Dia]], TableRest, Table).
+
+initialize_days_list([],_,_,_,_).
+initialize_days_list([H|T], N, NProfessores, FdSetDeProfs, ProfAvailable):-
+	H = [N,Hora,Prof],
+	Prof #= -1 #<=> Hora #= 0,
+	findall(ProfA, member( [ProfA, N], ProfAvailable), ListaDeProfsA),
+	list_to_fdset(ListaDeProfsA, FdSetDisponivel),
+	fdset_intersection(FdSetDisponivel, FdSetDeProfs, FdSetFinal),
+	Prof in_set FdSetFinal,
+	Hora in {0} \/{2},
+	N1 is N+1,
+	initialize_days_list(T, N1, NProfessores, FdSetDeProfs, ProfAvailable).
+
+initialize_solution([], _,_,_,_).
+initialize_solution([H|T], Lingua, NProfessores, Professores, ProfAvailable):-
+	H = Lingua-Lista,
+	length(Lista, 5),
+	professoresQueDaoLingua(Lingua, Professores, ListaDeProfs),
+	list_to_fdset([-1|ListaDeProfs], FdSetDeProfs),
+	initialize_days_list(Lista, 0, NProfessores,FdSetDeProfs,ProfAvailable),
+	Lingua1 is Lingua+1,
+	initialize_solution(T, Lingua1, NProfessores, Professores, ProfAvailable).
+
+	restrict_course_times([]).
+	restrict_course_times([_-[[_,H1,_],[_,H2,_],[_,H3,_],[_,H4,_],[_,H5,_]]|T]):-
+		sum([H1,H2,H3,H4,H5],#=<,4),
+		sum([H1,H2,H3,H4,H5],#\=,1),
+		sum([H1,H2,H3,H4,H5],#>=,0),
+		restrict_course_times(T).
+
+labelAll(H,P,Cost):- append(H, P, List), labeling([up,bisect,ff,timeout(60000,),
+maximize(Cost)], List).
+
+restrict_teacher_hours(_,_,[], _).
+restrict_teacher_hours(Horas, ProfessoresDominio, [[Indice,_,_]|T], NProfs):-
+	make_prof_table(Indice, Table, NProfs),!,
+	append(Table, [[-1, 0]], NTable),
+	restrict_teacher(NTable, ProfessoresDominio, Rest),
+	sum_product(Horas, Rest, Sum),
+	Sum #=< 15,
+	restrict_teacher_hours(Horas, ProfessoresDominio, T, NProfs).
+
+restrict_teacher(_, [], []).
+restrict_teacher(Table, [P|Ps], [Rest|T]):-
+	table([[P,Rest]], Table),
+	restrict_teacher(Table, Ps, T).
+
+sum_product([],[],0).
+sum_product([X|Xs],[Y|Ys], Sum):-
+	sum_product(Xs,Ys,SumRest),
+	Sum #= X*Y + SumRest.
+
+make_prof_table(Indice, Lista, NProfs):-
+	Length is NProfs+1,
+	length(Lista, Length),
+	iterate_prof_table(Lista, NProfs, Indice).
+
+iterate_prof_table(_,-1,_).
+iterate_prof_table(Lista, NProfs, NProfs):- !, nth0(NProfs, Lista, [NProfs, 1]),
+ Next is NProfs-1, iterate_prof_table(Lista, Next, NProfs).
+iterate_prof_table(Lista, NProfs, Indice):- nth0(NProfs, Lista, [NProfs,0]), 
+Next is NProfs-1, iterate_prof_table(Lista, Next, Indice).
+
+generate_domain_day_list([],[],[]).
+generate_domain_day_list([[_,Hora,Prof]|Dias], H, P):-
+	generate_domain_day_list(Dias, HRest, PRest),
+	append([Hora], HRest, H),
+	append([Prof], PRest, P).
+
+generate_variable_domain_list([],[],[]).
+generate_variable_domain_list([_-Dias|Solucoes], H, P):-
+	generate_domain_day_list(Dias, HThis, PThis),
+	generate_variable_domain_list(Solucoes, HRest, PRest),
+	append(HThis, HRest, H),
+	append(PThis, PRest, P).
+
+custo_nulo([], 0).
+custo_nulo(TabelaCustos, NLinguas):-
+	NLinguas \= 0,
+	N is NLinguas-1,
+	custo_nulo(TabelaCustosR, N),
+	append([[-1,N,0]], TabelaCustosR, TabelaCustos).
+
+escola_de_linguas(Professores, Candidaturas, Linguas, Rooms, Lucro, Solucao):-
+	length(Professores, NProfessores),
+	length(Linguas, NLinguas),
+	length(Solucao, NLinguas),
+	generate_professor_availability(Professores, ProfAvailable),
+	cria_tabela_custos(Professores, TabelaCustosReais, []),
+	custo_nulo(CustoNulo, NLinguas),
+	append(CustoNulo, TabelaCustosReais, TabelaCustos),
+	initialize_solution(Solucao, 0, NProfessores, Professores,ProfAvailable),
+	generate_days_prof_table(Solucao, DayProf),
+	restrict_rooms(DayProf,Rooms, NLinguas),
+	generate_variable_domain_list(Solucao, H, P),
+	restrict_teacher_hours(H, P, Professores, NProfessores),
+	restrict_course_times(Solucao),
+	make_profit(Solucao, Linguas, Professores, TabelaCustos, Candidaturas, Lucro),
+	labelAll(H, P, Lucro),
+	printSolution(Solucao),
+	 write('O lucro total obtido e de '), write(Lucro),  nl,
+	fd_statistics.
+
+cria_tabela_custos([], TabelaCustos, TabelaCustos).
+cria_tabela_custos([Professor|Professores], TabelaCustos, Acum):-
+	Professor = [Indice,Linguas,_],
+	cria_tabela_linguas(Linguas, Indice, TabelaLinguas, []),
+	append(Acum, TabelaLinguas, NovaTabela),
+	cria_tabela_custos(Professores, TabelaCustos, NovaTabela).
+
+cria_tabela_linguas([], _, T, T).
+cria_tabela_linguas([LinguaCusto|LinguasCustos], Indice, TabelaLinguas, Acum):-
+	LinguaCusto = IndiceL-Custo,
+	append(Acum, [[Indice,IndiceL,Custo]], NovaTabela),
+	cria_tabela_linguas(LinguasCustos, Indice, TabelaLinguas, NovaTabela).
+
+make_profit([], _, _, _,_, 0).
+make_profit([Solucao|Solucoes], Linguas, Professores,TabelaCustos, Candidaturas, Lucro):-
+	make_profit_language(Solucao, Linguas, Professores, TabelaCustos, Candidaturas
+	, LucroLingua),
+	make_profit(Solucoes, Linguas, Professores, TabelaCustos, Candidaturas, LucroResto),
+	Lucro #= LucroResto+LucroLingua.
+
+make_profit_language(Lingua, Linguas, Professores, TabelaCustos, Candidaturas,
+ LucroLingua):-
+	Lingua = IndiceL-Dias,
+	make_profit_day(Dias, Linguas, Professores, TabelaCustos, Candidaturas, IndiceL, 
+	LucroLingua).
+make_profit_day([], _, _,_, _, _, 0).
+make_profit_day([Dia|OutrosDias], Linguas, Professores, TabelaCustos, Candidaturas, 
+IndiceL, LucroDia):-
+	Dia = [_, Horas, Professor],
+	table([[Professor, IndiceL, Remuneracao]],TabelaCustos),
+	Prejuizo #= Remuneracao * Horas,
+	member(IndiceL-Preco, Linguas),
+	member(IndiceL-NumeroCandidatos, Candidaturas),
+
+	Inscritos is min(NumeroCandidatos, 15),
+	Lucro #= Preco*Inscritos*Horas,
+
+	LucroDiaAtual #= (Lucro-Prejuizo),
+	make_profit_day(OutrosDias, Linguas, Professores,TabelaCustos, Candidaturas, IndiceL
+	, LucroRest),
+	LucroDia #= LucroRest + LucroDiaAtual.
+
+generate_days_prof_table([], [[],[],[],[],[]]).
+generate_days_prof_table([_-[[_,_,Prof1],[_,_,Prof2],[_,_,Prof3],[_,_,Prof4],
+[_,_,Prof5]]|Solucoes], [Dia1, Dia2, Dia3, Dia4, Dia5]):-
+	generate_days_prof_table(Solucoes, [Dia1R, Dia2R, Dia3R, Dia4R, Dia5R]),
+	append([Prof1], Dia1R, Dia1),
+	append([Prof2], Dia2R, Dia2),
+	append([Prof3], Dia3R, Dia3),
+	append([Prof4], Dia4R, Dia4),
+	append([Prof5], Dia5R, Dia5).
+
+restrict_rooms([],_,_).
+restrict_rooms([Dia|Dias], Rooms, NLinguas):-
+	AulasVazias is NLinguas - Rooms,
+	count(-1, Dia, #>=, AulasVazias),
+	restrict_rooms(Dias, Rooms, NLinguas).
+
+printSolution([]).
+printSolution([Indice-Dias|Solutions]):-
+	nl, write('Language '), write(Indice), write(' Time Table'),nl, write('_______________'),
+	nl,
+	printLanguage(Indice, Dias),
+	printSolution(Solutions).
+
+dia(0, 'Monday').
+dia(1, 'Tuesday').
+dia(2, 'Wednesday').
+dia(3, 'Thursday').
+dia(4, 'Friday').
+
+printLanguage(_, []).
+printLanguage(_, [[_,0,-1]|Dias]):-printLanguage(Indice, Dias).
+printLanguage(Indice, [ [Dia,Horas,Prof]|Dias]):-
+	dia(Dia, DiaDaSemana),
+	write(DiaDaSemana), write(' - '), write(Horas), write(' hours with professor '), 
+	write(Prof), nl,
+	printLanguage(Indice, Dias).
+
+\end{verbatim}
+
 \end{document}

### 2015-12-13 proj2/algumasconsideracoessobreoproblemaemestudo.pl

@@ -32,9 +32,6 @@ professoresQueDaoLingua(Lingua,  [ [_, _,_] | OutrosProfs], Lista):-
 	professoresQueDaoLingua(Lingua, OutrosProfs, Lista).
 
 
-testeProfs(Sol):-
-	professoresQueDaoLingua(5, [[0, [0-1,1-2],_], [1, [0-2],_], [2, [3-2],_]], Sol).
-
 generate_professor_availability([], []).
 generate_professor_availability([[Indice,_,Dias]|Professors], Table):-
 	generate_professor_availability(Dias,Indice,TableThis),
@@ -156,20 +153,6 @@ escola_de_linguas(Professores, Candidaturas, Linguas, Rooms, Lucro, Solucao):-
 	printSolution(Solucao),
 	 write('O lucro total obtido e de '), write(Lucro),  nl,
 	fd_statistics.
-
-
-
-
-
-teste:-
-	statistics(runtime, [T0| _]),
-	%escola_de_linguas([[0, [0-0,1-0,3-0,4-0,5-0,6-0],[0,1,2,3,4]], [1, [0-0,2-0],[0,1,2,3,4]],[2, [0-0,3-0,4-0,5-0,6-0],[0,1,2,3,4]],[3, [0-0,1-0,3-0,4-0,5-0,6-0],[0,1,2,3,4]]], [0-10,1-10,2-10,3-10,4-10,5-10,6-10], [0-1,1-1,2-1,3-1,4-1,5-1,6-1], 40,Lucro, Solucao),
-	escola_de_linguas([[0, [0-1, 1-1,2-1,3-0,4-0],[0,1,4,5,6]], [1, [0-1, 1-1,2-1],[0,4]], [2, [0-1, 1-1,3-1],[0,3]] ],
-	[0-15, 1-10,2-10,3-10,4-10,5-10],
-	[0-1,1-1,2-1,3-10,4-10,5-10], 4,_, _),
-	statistics(runtime, [T1|_]),
-	T is T1 - T0,
-	format('O resultado foi obtido em ~3d segundos.~n', [T]).
 
 
 cria_tabela_custos([], TabelaCustos, TabelaCustos).

### 2015-12-13 proj2/doc/article.tex
Relatorio acabado

--- 2015-12-13 proj2/doc/article.tex
@@
 \usepackage[utf8]{inputenc}
+\usepackage{graphicx}
+\usepackage{float}
+\usepackage{placeins}
 %
@@
 %
-ISTO TA A FICAR APERTADO D PAGS, SE EXCEDER TIRA-SE OS VERBATIMS TODOS E OS PROFS PODEM IR LER AOS ANEXOS Q SE FODEM
+%ISTO TA A FICAR APERTADO D PAGS, SE EXCEDER TIRA-SE OS VERBATIMS TODOS E OS PROFS PODEM IR LER AOS ANEXOS Q SE FODEM
 \section{Introduction}
@@
 
-\subsection{Restraints}
+\subsection{Constraints}
 % FALTA BOTAR AQUI CENAS
-The restraints for this particular problem are listed in section 2.
-As such this section will focus mainly on restrictions' implementation aspects.
-All restricitions concerning monetary values are approached during the profit calculating phase of the solution, which is further explained in the next section of this paper.
-For the restriction concerning the number of hours of class a certain course can have per week, the program will gather the values for the hours taught each day and say that their sum must be grater than or equal to 2 and lesser than or equal to 4.
+The constraints for this particular problem are listed in section 2.
+As such this section will focus mainly on constraints' implementation aspects.
+All constraints concerning monetary values are approached during the profit calculating phase of the solution, which is further explained in the next section of this paper.
+For the constraint concerning the number of hours of class a certain course can have per week, the program will gather the values for the hours taught each day and say that their sum must be grater than or equal to 2 and lesser than or equal to 4.
 
@@
 It is then said that the sum of these variables cannot exceed 15 (the code concerning this can be viewed in the annex, and will not be replicated here for reasons of spatial economy).
+
+In the interest of reducing the domain and the search space to feasible levels the authors implemented a further constraint that limits each class to only having a 2 hour duration.
+This means that each course can have either one 2 hour class or 2 2 hour classes.
 
@@
 \begin{verbatim}
-labeling([ff,enum,down,maximize(Profit)], Flat).
+labeling([ff,leftmost,down,maximize(Profit)], Flat).
 \end{verbatim}
@@
 \subsection{Search Strategy}
- NUM SEI OQ HEI-DE POR AQUI
+Establishing an optimal search strategy for this problem presents difficulties due to the large variety of valid inputs possibly benefitting from a variety of different options.
+The search strategies selected are presented with a brief rationale for their choice.
+The tests run on each set of options are show on table \ref{table:speedup}.
+
+For choice making the \textit{step} option was selected over the common \textit{bisect} option.
+It allows for a speedier pruning of the domain space because, in conjunction with the \textit{down} option it leads to higher profit solutions as the first solutions.
+The discovery of high profit solutions early in the branch-and-bound algorithm allows for larger prunings to be made on the search tree.
+This results in marginally faster times when compared to \textit{bisect}.
+
+Given the goals of maximizing the cost being closely tied with maximizing the number of hours taught at the school the obvious choice of searching though domains in descendant order with the option \textit{down} was made.
+This resulted in significant speedups when compared to the default option \textit{up}.
+
+\begin{table}[]
+\caption{6 professors, 5 languages, 4 classrooms test}
+\label{table:speedup}
+\centering
+\begin{tabular}{l|l|l|}
+\cline{2-3}
+                                      & time(s) & speedup \\ \hline
+\multicolumn{1}{|l|}{up,bisect,ff}    & 520.052 & 1x      \\ \hline
+\multicolumn{1}{|l|}{down,bisect,ff,} & 20.725  & 25.09x  \\ \hline
+\multicolumn{1}{|l|}{down,step,ff}  & 21.349  & 25.55x  \\ \hline
+\end{tabular}
+
+\end{table}
 
@@
  \section{Results}
- %BOTA GRAFICOS E TAL
+The following graphics plot the performance results obtained with the modeling and constraints chosen, when increasing the number of classrooms, professors and languages being taught.
+All other variables are not changing in each series, to provide a accurate representation of the performance of the program.
 
+\begin{figure}[h]
+  \centering
+    \includegraphics[width=1\textwidth]{professor.png}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+    \includegraphics[width=1\textwidth]{classrooms.png}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+    \includegraphics[width=1\textwidth]{language.png}
+\end{figure}
+
+\FloatBarrier
  \section{Conclusions and Future Work}
@@
 Indeed, the implemented solution shows that the use of this technology allows for a relatively simple implementation of a solution, when compared to possible implementations using other technologies.
+
+It is possible to observe the effect on performance of the establishment of constraints, which works somewhat conterintuitively in that the imposition of further constraints can be extremely beneficial for performance, achieving sometimes surprising speedups by imposing constraints.
+One should however note that some constraints, particularly those that are not global, can also significantly degrade performance.
+The use of reification in particular, which the authors attempted in one of the first implementations, proved to be too costly despite the advantages in code simplicity it provided.
+It was thus replaced with a global constraint.
+
+The solution now presented suffers from a significant degradation in performance as the problem size expands.
+This is partly due to the intrinsic characteristics of the problem that mandates the use of multiple decision variables with  relatively large domains, compounded by the fact that each decision made by the solver can have a relatively modest effect on the size of the remaining domains when it propagates.
+However, the modelization chosen is also not optimal as it allows, within the constraints of the problem, for multiple, equivalent, solutions that while valid and distinct do not represent any sort of improvement or even to relevant alternatives.
+For example, professors who can give classes in a large range of days can provoke an exponential growth of the solution space, even when constrained to the optimal solutions, which the solver is forced to explore.
+Relevant future work includes improvements to the modelization of the problem and possibly the implementation of a heuristic of selection of domain variables to allow for quicker constraint of the domains and possibly discarding exploration of useless, equivalent paths in the search tree.
 %aspetos a melhorar???
